---
title: "Unified QC Report"
author: ""
output:
  html_document:
    code-fold: true
    self-contained: true
    toc: true
execute:
  echo: false
---

This interactive notebook generates a summary of QC statistics to assess the success of a **PfPHAST** or **MADHatTeR (MH)** run.

::: {.callout-note collapse="true"}
## Pipeline Mode Configuration

This unified QC report supports both PfPHAST and MADHatTeR (MH) pipelines.

**Set your pipeline mode in the configuration section below.**

### Key Differences Between Modes

| Feature | PfPHAST | MADHatTeR (MH) |
|---------|---------|----------------|
| Pool Structure | 60 targets in 2 pools (M1, M2)* | 276 targets in 2 pools (pool 1, pool 2) |
| Pool Assignment | M1.1 + M1.addon → pool 1, M2.1 → pool 2 | First number after letter: D1/R1.x → pool 1, R2 → pool 2 |
| Reaction Extraction | From pool file | From locus name (first digit after letter) |
| Add-on Handling | Optional: M1.addon can be excluded from QC | N/A |
| Polyclonality | 3 categories (unified) | 3 categories (unified) |
| WSAF Filtering | Optional via filter_by_wsaf flag | Optional via filter_by_wsaf flag |

*M1.addon loci can be excluded from QC calculations by setting `use_addons = FALSE`

### Unified Locus Exclusion (Both Modes)
**Common exclusions (both modes):**
- **5 species loci**: Non-P. falciparum species markers (ending in -1AB)
- **9 long amplicons**: Amplicons >275bp that tend to underperform (various suffixes)

**Mode-specific exclusions:**
- **PfPHAST**: M1.addon loci (if `use_addons = FALSE`)
- **MH**: R1.1-exclusive loci (loci ONLY in R1.1 primer pool, which is not used in practice)

**Important: Excluded loci are filtered only for QC calculations, NOT from the final allele_data output.** This ensures that:
- QC metrics (reprep/repool, success rates) reflect only relevant loci for the assay configuration
- Final filtered allele data includes ALL loci (including excluded ones) meeting read/AF thresholds
- Downstream analyses have access to complete allele information including species markers
:::

::: {.callout-note collapse="true"}
## Required Inputs

To proceed, you must provide the **results directory** from the Mad4hatter pipeline, which should include the following files:

* **sample_coverage.txt**
* **amplicon_coverage.txt**
* **allele_data.txt**

Additionally, a **sample manifest** is required. This file must contain the following fields:

* **SampleID** – Unique identifier for each sample.
* **SampleType** – Specifies whether the entry is a **sample**, **positive** control, or **negative** control.
* **Batch** – Identifies a group of samples processed simultaneously by the same individual.
* **Column** – The well column where the sample was placed in the plate.
* **Row** – The well row where the sample was placed in the plate.
* **Parasitemia** – The qPCR value for the sample.
:::

::: {.callout-note collapse="true"}
## QC Summaries
The notebook provides the following analyses:

* **Plate Layout**: Location of samples and controls on the plate
* **Primer Dimer Content**: Input reads vs. % attributed to primer dimers
* **Balancing Across Batches**: Swarm plot of reads output by the pipeline by batch
* **Control Summary**, including:
    * Polyclonality of Positive Controls
    * Read Summary for Negative Controls
    * Contamination Maps for Negative Controls
* **Successul Amplification** summary, including:
    * Sample Reads vs. Number of Successfully Amplified Loci
    * Parasitemia vs. Number of Successfully Amplified Loci
* **Amplification Plate Maps**, including:
    * Reads Heatmap
    * Successful Amplification Heatmap


Note: A locus is considered successfully amplified if it has more than `read_threshold` reads, where `read_threshold` is a threshold that can be set below.
:::


## Setup

::: {.callout-note collapse="true"}
## Configuration and Data Loading

```{r}
# Running mode - if running interactively, for example in Rstudio, set this to FALSE
render_mode <- TRUE
```

```{r}
#####################################################
## USER CONFIGURATION - MODIFY THESE VALUES
#####################################################


# Set working directory (where pool files and output directories are located)
working_dir <- "~/Documents/PfPHAST_vs_MH_update_QC/"
setwd(working_dir)

# PIPELINE MODE: Set to "PfPHAST" or "MH" for MADHatTeR
pipeline_mode <- "PfPHAST"

# ADD-ON USAGE (PfPHAST only): Set to TRUE if M1.addon loci were used, FALSE to exclude them from QC
use_addons <- FALSE

# Input path to results directory from Mad4hatter pipeline
results_dir <- "~/Library/CloudStorage/Box-Box/IMMRSE-U Study/CNV_analysis/runs/IMMRSE Phase 5/NextSeq/Results-2025-11-20-NextSeq-015//"

# Input path to manifest
manifest_file <- "~/Library/CloudStorage/Box-Box/IMMRSE-U Study/CNV_analysis/manifests/IMMRSE Phase 5/NextSeq015_manifest.csv"

# Output file paths (absolute paths to ensure correct location)
reprep_repool_summary_path <- file.path(working_dir, "Output/QC/NextSeq015/reprep_repool_summary.csv")
polyclonal_information_path <- file.path(working_dir, "Output/QC/NextSeq015/polyclonal_controls.csv")
neg_control_information_path <- file.path(working_dir, "Output/QC/NextSeq015/negative_control_contamination.csv")
filtered_allele_data_path <- file.path(working_dir, "Output/QC/NextSeq015/allele_data_filtered.txt")
missing_loci_report_path <- file.path(working_dir, "Output/QC/NextSeq015/missing_loci_report.csv")
samples_to_repool_path <- file.path(working_dir, "Output/QC/NextSeq015/samples_to_repool.csv")
samples_to_reprep_path <- file.path(working_dir, "Output/QC/NextSeq015/samples_to_reprep.csv")
missing_samples_report_path <- file.path(working_dir, "Output/QC/NextSeq015/missing_samples_report.csv")

```

```{r}
# Validate pipeline mode
if (!pipeline_mode %in% c("PfPHAST", "MH")) {
  stop("Invalid pipeline_mode. Must be either 'PfPHAST' or 'MH'.")
}
cat("Pipeline Mode:", pipeline_mode, "\n")
```

```{r}
#####################################################
## QC THRESHOLDS AND FILTERS - ADJUST AS NEEDED
#####################################################

# If TRUE, trims extra sequencing-related information from SampleID to match the manifest format
standardise_sampleID <- TRUE

# Input a threshold for the reads needed to classify a locus as having amplified successfully
read_threshold <- 100

# Input a minimum Reads per ASV used as a filter for the positive control plot and final allele data
read_filter <- 0

# Minimum Allele Frequency per ASV used as a filter for positive control plot and final allele data
af_filter <- 0.01

# Apply WSAF (within-sample allele frequency) filter to final allele data output
# Set to FALSE to output all alleles regardless of WSAF (will still filter by read count)
filter_by_wsaf <- FALSE

# Read threshold to output locus information for negative control
negative_control_read_threshold <- 50

# Proportion of targets successfully amplified; samples below this threshold require re-prep
reprep_threshold <- 0.5

# Proportion of targets successfully amplified; samples below this threshold require re-pool
repool_threshold <- 0.75

# Allele ID column: Set to "PseudoCIGAR" to collapse to unique PseudoCIGARs, or "ASV" for ASVs
allele_col <- "PseudoCIGAR"
```

```{r}
# Print summary of user inputs
cat("============================================\n")
cat("USER CONFIGURATION SUMMARY\n")
cat("============================================\n")
cat("Pipeline Mode:", pipeline_mode, "\n")
if (pipeline_mode == "PfPHAST") {
  cat("Use Add-ons:", use_addons, "\n")
}
cat("Results Directory:", results_dir, "\n")
cat("Manifest File:", manifest_file, "\n")
cat("Read Threshold for Successful Amplification:", read_threshold, "\n")
cat("Minimum Reads per ASV (Positive Control Filter):", read_filter, "\n")
cat("Minimum Allele Frequency per ASV (Positive Control Filter):", af_filter, "\n")
cat("Filter Final Allele Data by WSAF:", filter_by_wsaf, "\n")
cat("Allele Column:", allele_col, "\n")
cat("============================================\n")
```

```{r}
# Create file paths (standard Mad4hatter output files)
sample_coverage_file <- file.path(results_dir, "sample_coverage.txt")
amplicon_coverage_file <- file.path(results_dir, "amplicon_coverage.txt")
allele_data_path <- file.path(results_dir, "allele_data.txt")
```

```{r}
# Load necessary libraries
library(ggplot2)
library(plotly, quietly = TRUE, warn.conflicts = FALSE)
library(dplyr, quietly = TRUE, warn.conflicts = FALSE)
library(tidyr)
library(ggbeeswarm)
library(stringr)
library(tibble)
library(purrr)
library(kableExtra, quietly = TRUE, warn.conflicts = FALSE)
library(gt)
```

```{r}
# UNIFIED EXCLUSION LISTS (used by both PfPHAST and MH modes)
# These loci are excluded from QC calculations but included in final output

# Species loci (non-P. falciparum markers, ending in -1AB)
species_loci <- c(
  'Pf3D7_13_v3-1041593-1041860-1AB',
  'PmUG01_12_v1-1397996-1398245-1AB',
  'PocGH01_12_v1-1106456-1106697-1AB',
  'PvP01_12_v1-1184983-1185208-1AB',
  'PKNH_12_v2-198869-199113-1AB'
)

# Long amplicons (>275bp, various suffixes)
long_loci <- c(
  'Pf3D7_14_v3-854128-854427-1B',
  'Pf3D7_11_v3-1294284-1294580-1B',
  'Pf3D7_13_v3-1465010-1465307-1B',
  'Pf3D7_13_v3-2841390-2841688-1B2',
  'Pf3D7_13_v3-2844334-2844629-1B2',
  'Pf3D7_02_v3-320652-320949-2',
  'Pf3D7_03_v3-240957-241256-2',
  'Pf3D7_05_v3-615379-615678-2',
  'Pf3D7_06_v3-857454-857753-2'
)

# R1.1-exclusive loci will be identified dynamically for MH mode

# Mode Configuration Object
config <- list(
  PfPHAST = list(
    pool_file = file.path(working_dir, "Data/Pools_PfPHAST_Pools.csv"),
    reaction_source = "pool_file"
  ),
  MH = list(
    pool_file = file.path(working_dir, "Data/Pools - MADHatTeR Pools content.csv"),
    reaction_source = "pool_file"
  )
)
active_config <- config[[pipeline_mode]]
```

```{r}
manifest <- read.csv(manifest_file) %>% mutate(Row = toupper(Row)) # Ensure Row is capitalised
sample_coverage <- read.csv(sample_coverage_file, sep="\t")
allele_data <- read.csv(allele_data_path, sep='\t')
```

```{r}
# Load coverage per amplicon
amplicon_coverage <- read.delim(amplicon_coverage_file)
```

```{r}
# Standardize SampleID if requested
amplicon_coverage <- amplicon_coverage %>%
  mutate(SampleID = if (standardise_sampleID) SampleID %>%
           sub("_L001$", "", .) %>%
           word(., start = 1, end = -2, sep = "_") else SampleID)
allele_data <- allele_data %>%
  mutate(SampleID = if (standardise_sampleID) SampleID %>%
           sub("_L001$", "", .) %>%
           word(., start = 1, end = -2, sep = "_") else SampleID)
```
:::

::: {.callout-note collapse="true"}
## Section A: Locus Filtering

Here we filter loci for QC calculations only. The excluded loci will still be included in the final allele_data output.

```{r}
# Store original allele_data (unfiltered) for final output
allele_data_original <- allele_data

# UNIFIED EXCLUSION: Both modes use the same hardcoded list
excluded_loci <- c(species_loci, long_loci)

# Additional exclusion for PfPHAST mode: M1.addon loci if use_addons is FALSE
addon_loci_excluded <- c()
if (pipeline_mode == "PfPHAST" && !use_addons) {
  # Need to load pool file temporarily to identify M1.addon loci
  pool_file_path <- file.path(working_dir, "Data/Pools_PfPHAST_Pools.csv")
  if (file.exists(pool_file_path)) {
    pool_temp <- read.csv(pool_file_path) %>%
      rename(Locus = Target)
    addon_loci_excluded <- pool_temp %>%
      filter(grepl("^M1\\.addon", Pools)) %>%
      pull(Locus)
    excluded_loci <- c(excluded_loci, addon_loci_excluded)
    cat("PfPHAST Mode with Add-ons DISABLED: Excluding", length(addon_loci_excluded), "M1.addon loci from QC\n")
  }
}

# Additional exclusion for MH mode: R1.1-exclusive loci (not used in R1.2 primer pool)
r11_only_loci_excluded <- c()
if (pipeline_mode == "MH") {
  # Load pool file to identify R1.1-exclusive loci
  if (file.exists(active_config$pool_file)) {
    pool_temp <- read.csv(active_config$pool_file) %>%
      rename(Locus = Target)
    r11_only_loci_excluded <- pool_temp %>%
      filter(Pools == "R1.1") %>%  # Exactly "R1.1", not "R1.1;R1.2" or other combinations
      pull(Locus)
    excluded_loci <- c(excluded_loci, r11_only_loci_excluded)
    cat("MH Mode: Excluding", length(r11_only_loci_excluded), "R1.1-exclusive loci (R1.1 primers not used)\n")
  }
}

cat(pipeline_mode, "Mode: Excluding", length(excluded_loci), "loci for QC calculations\n")
cat("  -", length(species_loci), "species loci (non-P. falciparum markers)\n")
cat("  -", length(long_loci), "long amplicons (>275bp)\n")
if (length(addon_loci_excluded) > 0) {
  cat("  -", length(addon_loci_excluded), "M1.addon loci (add-ons not used)\n")
}
if (length(r11_only_loci_excluded) > 0) {
  cat("  -", length(r11_only_loci_excluded), "R1.1-exclusive loci (R1.1 primers not used)\n")
}
cat("\nExcluded species loci:\n")
for (locus in species_loci) {
  cat("  ", locus, "\n")
}
cat("\nExcluded long loci:\n")
for (locus in long_loci) {
  cat("  ", locus, "\n")
}
if (length(addon_loci_excluded) > 0) {
  cat("\nExcluded M1.addon loci (first 10):\n")
  for (locus in head(addon_loci_excluded, 10)) {
    cat("  ", locus, "\n")
  }
  if (length(addon_loci_excluded) > 10) {
    cat("  ... and", length(addon_loci_excluded) - 10, "more\n")
  }
}
if (length(r11_only_loci_excluded) > 0) {
  cat("\nExcluded R1.1-exclusive loci (first 10):\n")
  for (locus in head(r11_only_loci_excluded, 10)) {
    cat("  ", locus, "\n")
  }
  if (length(r11_only_loci_excluded) > 10) {
    cat("  ... and", length(r11_only_loci_excluded) - 10, "more\n")
  }
}

# Create filtered datasets for QC calculations only
amplicon_coverage_qc <- amplicon_coverage %>% filter(!Locus %in% excluded_loci)
allele_data_qc <- allele_data %>% filter(!Locus %in% excluded_loci)

cat("\nLocus counts:\n")
cat("  Total loci in amplicon_coverage:", n_distinct(amplicon_coverage$Locus), "\n")
cat("  Loci used for QC (after filtering):", n_distinct(amplicon_coverage_qc$Locus), "\n")
cat("  Excluded loci will still appear in final allele_data_filtered.txt output\n")
```
:::

::: {.callout-note collapse="true"}
## Section B: Pool File Loading & Reaction Assignment

```{r}
# Initialize pool_info as NULL
pool_info <- NULL

if (active_config$reaction_source == "pool_file") {
  # Both modes: load pool file and extract reactions
  if (file.exists(active_config$pool_file)) {
    pool_info <- read.csv(active_config$pool_file) %>%
      rename(Locus = Target)

    if (pipeline_mode == "PfPHAST") {
      # PfPHAST: M1.1 and M1.addon -> pool "1", M2.1 -> pool "2"
      pool_info <- pool_info %>%
        mutate(reaction = case_when(
          grepl("^M1\\.", Pools) ~ "1",
          grepl("^M2\\.", Pools) ~ "2",
          TRUE ~ NA_character_
        ))
    } else {
      # MH: Extract first digit after letter from Pools column
      # D1 -> "1", R1.1/R1.2 -> "1", R2.1/R2.2 -> "2"
      pool_info <- pool_info %>%
        mutate(reaction = str_extract(Pools, "(?<=[A-Z])\\d+(?=[\\.A-Z]|$)"))
    }

    # Use amplicon_coverage_qc for loci counting (not allele_data)
    nloci_table <- amplicon_coverage_qc %>%
      distinct(Locus) %>%
      left_join(pool_info %>% select(Locus, reaction), by = "Locus") %>%
      filter(!is.na(reaction)) %>%
      count(reaction) %>%
      rename(nreactionloci = n)

    cat("Loaded pool file:", active_config$pool_file, "\n")
    cat("Pool assignment:",
        if(pipeline_mode == "PfPHAST") "M1.x -> pool 1, M2.x -> pool 2"
        else "First digit after letter: D1/R1.x -> pool 1, R2.x -> pool 2", "\n")
    cat("Pools found:", paste(nloci_table$reaction, collapse=", "), "\n")
    cat("Loci per pool:\n")
    print(nloci_table)
  } else {
    warning("Pool file not found:", active_config$pool_file)
    nloci_table <- data.frame(reaction = character(), nreactionloci = integer())
  }
} else {
  stop("Invalid reaction_source in config: '", active_config$reaction_source, "'")
}
```
:::

```{r}
merge_data <- function(manifest, sample_coverage, standardise_sampleID) {
    merged_data <- sample_coverage %>%
        mutate(SampleID = if (standardise_sampleID) SampleID %>%
               sub("_L001$", "", .) %>%
               word(., start = 1, end = -2, sep = "_") else SampleID) %>%
        mutate(Reads = as.numeric(Reads)) %>%
        replace_na(list(Reads = 0)) %>%
        pivot_wider(names_from = Stage, values_from = Reads) %>%
        inner_join(manifest, by = "SampleID")
    return(merged_data)
}
```

```{r}
# Reformat sample_coverage and add sample information from manifest
sample_coverage_with_manifest <- merge_data(manifest, sample_coverage, standardise_sampleID)
```

```{r}
# Add manifest to amplicon coverage AND add reaction column (COMBINED FOR ATOMICITY)
amplicon_coverage_with_manifest <- amplicon_coverage_qc %>%
    inner_join(manifest, by = c("SampleID"))

# Add reaction column from pool file (both modes)
if (!exists("pool_info") || is.null(pool_info)) {
  stop("ERROR: pool_info not loaded. Pool file is required for both PfPHAST and MH modes.\n",
       "Expected pool file: ", active_config$pool_file)
}

amplicon_coverage_with_manifest <- amplicon_coverage_with_manifest %>%
  left_join(pool_info %>% select(Locus, reaction), by = "Locus")

cat("Added reaction column from pool file\n")

# Verify reaction column was added successfully
if (!"reaction" %in% colnames(amplicon_coverage_with_manifest)) {
  stop("CRITICAL ERROR: Failed to add 'reaction' column to amplicon_coverage_with_manifest.\n",
       "This will cause downstream joins to fail.")
}

# Report success
cat("✓ Successfully added reaction column\n")
cat("  Unique reactions found:", paste(sort(unique(amplicon_coverage_with_manifest$reaction[!is.na(amplicon_coverage_with_manifest$reaction)])), collapse=", "), "\n")
cat("  Rows with NA reaction:", sum(is.na(amplicon_coverage_with_manifest$reaction)), "\n")

# Report which loci have NA reactions (for debugging)
if (sum(is.na(amplicon_coverage_with_manifest$reaction)) > 0) {
  na_loci <- amplicon_coverage_with_manifest %>%
    filter(is.na(reaction)) %>%
    distinct(Locus) %>%
    pull(Locus)
  cat("  Loci with NA reactions:", paste(head(na_loci, 5), collapse=", "),
      ifelse(length(na_loci) > 5, paste0("... (", length(na_loci), " total)"), ""), "\n")
}
```

```{r}
# Create unfiltered version for negative control analysis (includes all loci)
amplicon_coverage_with_manifest_full <- amplicon_coverage %>%
    inner_join(manifest, by = c("SampleID"))

# Add reaction column to the unfiltered version (same logic)
amplicon_coverage_with_manifest_full <- amplicon_coverage_with_manifest_full %>%
  left_join(pool_info %>% select(Locus, reaction), by = "Locus")

cat("Created amplicon_coverage_with_manifest_full (includes all loci for negative control analysis)\n")
cat("  Filtered version loci:", n_distinct(amplicon_coverage_with_manifest$Locus), "\n")
cat("  Full version loci:", n_distinct(amplicon_coverage_with_manifest_full$Locus), "\n")
```

```{r}
# Create sample summary stats (uses QC-filtered amplicon_coverage_with_manifest)
summary_samples <- amplicon_coverage_with_manifest %>%
  # Join the nloci_table
  left_join(nloci_table, by = "reaction") %>%
  # Calculate number of loci with reads over threshold per reaction
  group_by(SampleID, Batch, SampleType, reaction, nreactionloci) %>%
  summarize(
    reads_per_reaction = sum(OutputPostprocessing),
    n_good_loci = sum(OutputPostprocessing > read_threshold),
    .groups = "drop"
  ) %>%
    group_by(SampleID, Batch, SampleType) %>%
      mutate(
        reads_per_sample = sum(reads_per_reaction),
      ) %>%
  ungroup() %>%
  mutate(prop_good_loci = n_good_loci / nreactionloci) %>%
  inner_join(manifest, by = join_by(SampleID, Batch, SampleType))
```

```{r}
# Plotting Setup
sample_colours <- c(
  "negative" = "red3",
  "positive" = "blue3",
  "sample" = "darkgrey",
  "empty" = "black"
)
```

::: {.callout-note collapse="true"}
## Section E: Reaction Color Assignment

```{r}
# Define reaction colors based on available reactions from BOTH filtered and unfiltered datasets
# This ensures colors are available for all plots (including negative controls which use full data)
reactions_filtered <- unique(amplicon_coverage_with_manifest$reaction)
reactions_full <- unique(amplicon_coverage_with_manifest_full$reaction)
reactions <- unique(c(reactions_filtered, reactions_full))
reactions <- reactions[!is.na(reactions)]

if (length(reactions) > 0) {
  reaction_colours <- setNames(
    scales::hue_pal()(length(reactions)),
    reactions
  )
  cat("Reaction colors assigned for:", paste(reactions, collapse=", "), "\n")
} else {
  # Fallback: create a default color palette
  reaction_colours <- c("1" = "#F8766D", "2" = "#00BFC4")
  cat("WARNING: No reactions found in data. Using default color palette.\n")
}
```
:::

```{r}
# Print summary of samples in files
cat("Sample Summary\n")
cat(length(unique(manifest$SampleID)), "samples in manifest", "\n")
cat(length(unique(allele_data$SampleID)), "samples in allele data file (unfiltered)", "\n")
cat(length(unique(allele_data_qc$SampleID)), "samples in allele data file (QC-filtered)", "\n")
cat(length(unique(sample_coverage$SampleID)), "samples in sample coverage file", "\n")
cat(length(unique(amplicon_coverage_qc$SampleID)), "samples in amplicon coverage file (QC-filtered)", "\n")
```

## Plate Layout

::: {.callout-tip collapse="true"}
## About this section
The plate layout provides a visual representation of sample organization, including the placement of positive and negative controls. This is allows verification that controls are positioned correctly to validate assay performance with later plots.
:::

```{r}
create_tabsetOfGgplotObjects <- function(ggplotObjectsList) {
  cat("\n\n::: {.panel-tabset}\n\n")
  iwalk(ggplotObjectsList, ~ {
    cat('## ', .y, '\n\n')
    print(.x)
    cat('\n\n')
  })
  cat(":::\n")
}
```

```{r}
create_plate_template <- function(summary_samples, batches, nrows=8, ncols=12) {
    quadrants <-
      expand.grid(
        Batch = batches,
        y = 1:nrows,
        Column = 1:ncols
      ) %>%
      mutate(
        ymin = y - 0.45,
        ymax = y + 0.45,
        xmin = Column - 0.45,
        xmax = Column + 0.45
      ) %>%
      mutate(Row = toupper(rev(letters[1:nrows])[y])) %>%
      # Join on the sample information to the plate
      left_join(
        summary_samples %>%
          select(Batch, Column, Row, SampleType) %>%
          distinct(),
        by = c("Batch", "Column", "Row" = "Row")
      )
    return(quadrants)
}
```

```{r}
quadrants <- create_plate_template(summary_samples, unique(summary_samples$Batch))
```

```{r}
plot_plate <- function(quadrants_batch, sample_colours, batch_name) {
    plate_layout <- ggplot(quadrants_batch) +
          # Tiles representing sample types, fill based on SampleType
          geom_tile(aes(x = Column, y = y, fill = SampleType), color = "white") +
          scale_y_continuous(breaks = quadrants_batch$y, labels = quadrants_batch$Row) +
          # Make sure labels are correct
          xlab("Column") +
          ylab("Row") +
          # Customizing the x-axis
          scale_x_continuous(breaks = 1:12, labels = as.character(1:12)) +
          # Quadrants as rectangles with borders
          geom_rect(
            data = quadrants_batch,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            color = "black", fill = NA, linewidth = 1
          ) +
          # Custom fill colors for each SampleType
          scale_fill_manual(
            values = sample_colours
          ) +
          # Theming
          theme_minimal() +
          theme(
            axis.text = element_text(size = 12),
            strip.text = element_text(size = 14, face = "bold"),
            panel.grid = element_blank(),
            plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
          ) +
        # Customizing the legend to add an outline for the colored squares
        guides(
          fill = guide_legend(
            override.aes = list(
              color = "black",  # Outline color for the squares
              size = 1          # Outline size for the squares
            )
          )
        ) +
        # Title for the plot
        ggtitle(paste0("Plate Map: ", batch_name)) +
        coord_fixed(ratio = 0.75)
    return(plate_layout)
}

plot_plate_layouts <- function(plate_data, sample_colours) {
    plate_maps = list()
    plate_data$SampleType[is.na(plate_data$SampleType)] <- "empty"
    for(b in unique(plate_data$Batch)){
      quadrants_batch = plate_data %>%
        filter(Batch == b)
      plate_maps[[b]] = plot_plate(quadrants_batch, sample_colours, b)
    }
    return(plate_maps)
}
```

```{r}
#| results: asis
#| echo: false
#| fig-column: screen-inset
plate_maps <- plot_plate_layouts(quadrants, sample_colours)
if (render_mode) {
  cat(create_tabsetOfGgplotObjects(plate_maps))
} else {
  for (plate in plate_maps) {
    print(plate)
  }
}
```

## Primer Dimer Content

::: {.callout-tip collapse="true"}
## About this section
Here we visualise the proportion of sequencing reads that are classified as primer dimers, which occur when primers anneal to each other instead of the target DNA. This plot is useful for assessing the efficiency of the amplification process, as high primer dimer levels can indicate suboptimal reaction conditions, reduced sequencing efficiency, and potential issues with sample quality or reagent performance.
:::

```{r}
# Function to generate the plot
generate_dimer_plot <- function(sample_coverage_with_manifest, sample_colours) {

    dimer_plot <- ggplot(data = sample_coverage_with_manifest %>%  arrange(desc(SampleType))) +
      # Plot points with different colors for each sample type
      geom_point(aes(x = Input + 0.9,
                     y = (1 - OutputPostprocessing / Input) * 100,
                     color = SampleType),
                 shape = 1,
                 alpha=0.8,
                 stroke = 1) +
    # Log scale for the x-axis
    scale_x_log10() +
    # Faceting by Batch column
    facet_wrap(~Batch) +
    # Adding labels and title
    ylab("% Dimers") +
    xlab("Input Reads") +
    ggtitle("Dimer Content")+
    # Color scale for SampleType
    scale_color_manual(values = sample_colours)
    print(dimer_plot)
}
```

```{r}
generate_dimer_plot(sample_coverage_with_manifest, sample_colours)
```

## Balancing Across Batches

::: {.callout-tip collapse="true"}
## About this section
Here we show the distribution of total reads per sample across different batches, helping to assess whether sequencing depth is consistent. This plot is useful when you have multiple batches in the same sequencing run to identify imbalances in sequencing, which can arise due to variations in library preparation, loading efficiency, or sequencing conditions.
:::

```{r}
generate_balancing_plot <- function(summary_samples, sample_colours) {
  ggplot() +
    ggbeeswarm::geom_quasirandom(
    data = summary_samples %>%
        select(SampleID, Batch, reads_per_sample, SampleType) %>%
        distinct(),
    aes(x = Batch, y = reads_per_sample + 0.9, color = SampleType)
    ) +
    scale_y_log10() +
    scale_color_manual(values = sample_colours) +
    ylab("Total Reads for Sample") +
    ggtitle("Balancing Across Batches") +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1)
    )
}
```

```{r}
generate_balancing_plot(summary_samples, sample_colours)
```

## Control Summary

### Polyclonality of Positive Controls

::: {.callout-tip collapse="true"}
## About this section
Here we inspect the positive controls, ensuring that the positive controls perform as expected without contamination or unwanted diversity. For example, if the control you included was monoclonal you would expect little to no loci reported as having more than one allele (polyclonal). In some cases you may allow for some level of false positive detection within a monoclonal control.
:::

```{r}
# Add on within sample allele frequency (use QC version)
allele_data_qc <- allele_data_qc %>%
  group_by(SampleID, Locus) %>%
  mutate(AlleleFreq = Reads / sum(Reads)) %>%
  ungroup()
```

```{r}
pos_control_allele_data <- allele_data_qc %>%
  inner_join(manifest %>% filter(SampleType == "positive"), by = "SampleID")
```

```{r}
if (allele_col == "PseudoCIGAR") {
  pos_control_data <- pos_control_allele_data %>%
      group_by(SampleID, Locus, PseudoCIGAR) %>%
      summarise(Reads = sum(Reads), AlleleFreq = sum(AlleleFreq), .groups = "drop")
} else {
  pos_control_data <- pos_control_allele_data
}
```

```{r}
# Apply filters and identify the number of alleles being called per locus
locus_summary <- pos_control_data %>%
  group_by(SampleID, Locus) %>%
  summarise(
    NumASVs_Meeting_Threshold = sum(Reads > read_filter & AlleleFreq > af_filter),
    .groups = "drop"
  )

# Unified 3-category polyclonality classification (both modes)
locus_summary <- locus_summary %>%
  mutate(
    Category = case_when(
      NumASVs_Meeting_Threshold == 1 ~ "Monoclonal",
      NumASVs_Meeting_Threshold == 2 ~ "2 Alleles",
      NumASVs_Meeting_Threshold > 2  ~ ">2 Alleles"
    )
  )
fill_colors <- c("Monoclonal" = "Gray", "2 Alleles" = "orangered2", ">2 Alleles" = "turquoise3")

# Count loci per sample for each category
result <- locus_summary %>%
  group_by(SampleID, Category) %>%
  summarise(Count = n(), .groups = "drop")

# Create ggplot
p <- ggplot(result, aes(x = SampleID, y = Count, fill = Category, text = paste("Count:", Count))) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Clonality of Loci for Positive Controls", x = "Sample ID", y = "Number of Loci") +
  theme_minimal() +
  scale_fill_manual(values = fill_colors) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)
  )

# Convert ggplot to interactive plot
ggplotly(p, tooltip = "text", width = 800, height = 500)
```

```{r}
# Filter data to keep only polyclonal loci (2 or more alleles)
polyclonal_information <- pos_control_data %>%
  inner_join(locus_summary, by = c("SampleID", "Locus")) %>%
  filter(Category %in% c("2 Alleles", ">2 Alleles"))

# Display as styled table
head(polyclonal_information) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Negative Control Contamination

#### Read Summary per Negative Control

::: {.callout-tip collapse="true"}
## About this section
In an ideal scenario, negative controls should have minimal or no amplification. If a negative control shows a high number of targets amplified with significant reads, it suggests potential contamination. By plotting the number of reads against the number of targets for each sample, any outliers or unexpected amplification in negative controls can be easily flagged.
:::

```{r}
#| fig-width: 12
#| fig-height: 8
# retrieve negative controls from amplicons.full (use unfiltered version to include all loci)
amplicons_negative = amplicon_coverage_with_manifest_full %>%
  filter(SampleType=="negative") %>%
  mutate(negative = paste0(Batch," Well: ",toupper(Row),Column))
```

```{r}
# Get negative controls grouped by batch
batches_with_negatives <- amplicons_negative %>%
  group_by(Batch) %>%
  summarize(
    n_controls = n_distinct(negative),
    .groups = "drop"
  ) %>%
  arrange(Batch)

cat("Negative controls by batch:\n")
print(batches_with_negatives)

n_batches <- nrow(batches_with_negatives)

# Define layout parameters
ncol_layout <- 3         # Number of columns in facet grid
base_width <- 12
base_height_per_row <- 2.5  # Height per row of facets
min_height <- 6
max_height <- 20  # Prevent excessively tall figures

cat("Total batches with negative controls:", n_batches, "\n")
```

```{r}
# Plotting helper function for negative control histograms
plot_negative_control_histogram <- function(data, reaction_colours, batch_name = NULL) {
  title_suffix <- if (!is.null(batch_name)) paste0(" - ", batch_name) else ""

  p <- ggplot(data) +
    geom_histogram(aes(x = OutputPostprocessing, fill = reaction)) +
    facet_wrap(~ negative, ncol = ncol_layout, scales = "free_y") +
    xlim(-ifelse(
      max(data$OutputPostprocessing) == 0,
      1e2,
      max(data$OutputPostprocessing)
    ) / 50,
    ifelse(
      max(data$OutputPostprocessing) == 0,
      1e2,
      max(data$OutputPostprocessing) + 10
    )) +
    xlab("Reads") +
    ylab("Number of targets") +
    ggtitle(paste0("Read Distribution per Negative Control", title_suffix)) +
    scale_fill_manual(values = reaction_colours) +
    theme_minimal(base_size = 14) +
    theme(
      axis.text = element_text(size = 12),
      strip.text = element_text(size = 11, face = "bold"),
      panel.spacing = unit(1, "lines")
    )

  return(p)
}
```

```{r}
#| results: asis
#| echo: false

if (n_batches == 1) {
  # Single batch - no tabs needed, just dynamic height
  batch_data <- amplicons_negative
  n_controls <- n_distinct(batch_data$negative)

  # Calculate height for this batch
  batch_height <- min(max_height, max(min_height,
                                       ceiling(n_controls / ncol_layout) * base_height_per_row))

  knitr::opts_chunk$set(fig.width = base_width, fig.height = batch_height)

  p <- plot_negative_control_histogram(batch_data, reaction_colours)
  print(p)

} else {
  # Multiple batches - create tabs for each batch
  cat("\n\n::: {.panel-tabset}\n\n")

  batch_list <- unique(amplicons_negative$Batch)

  for (batch in batch_list) {
    cat('## ', batch, '\n\n')

    # Filter data for this batch
    batch_data <- amplicons_negative %>%
      filter(Batch == batch)

    # Calculate number of controls in this batch
    n_controls_in_batch <- n_distinct(batch_data$negative)

    # Calculate height for this batch
    batch_height <- min(max_height, max(min_height,
                                         ceiling(n_controls_in_batch / ncol_layout) * base_height_per_row))

    # Set figure dimensions for this batch
    knitr::opts_chunk$set(fig.width = base_width, fig.height = batch_height)

    cat("Negative controls in", batch, ":", n_controls_in_batch, "\n\n")

    # Create and print plot
    p <- plot_negative_control_histogram(batch_data, reaction_colours, batch)
    print(p)

    cat('\n\n')
  }

  cat(":::\n\n")
}

# Reset chunk options to defaults
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
```

```{r}
neg_control_information <- amplicons_negative %>%
    filter(OutputPostprocessing > negative_control_read_threshold) %>%
    select('SampleID','Locus','OutputPostprocessing')
head(neg_control_information) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

#### Aggregated Reads per Target

::: {.callout-tip collapse="true"}
## About this section
This plot aggregates the total reads for each locus across all negative controls. In negative controls, if there is an unexpected spike in reads for specific loci, it could indicate contamination in the form of cross-sample contamination or environmental contamination. Analyzing these total summed reads helps pinpoint specific loci where contamination might have occurred, offering insights into which steps in the process might have introduced contaminants.
:::

```{r}
#| fig-width: 12
#| fig-height: 8
# Get reads for each locus, label reaction and extract chrom and start from locus name
summary_data <- amplicons_negative %>%
  group_by(reaction, Locus) %>%
  summarise(sum_reads = sum(Reads, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    Chromosome = sub("-.*", "", Locus),
    Start = as.numeric(sub(".*-(\\d+)-.*", "\\1", Locus))
  ) %>%
  arrange(Chromosome, Start) %>%
  mutate(Locus = factor(Locus, levels = unique(Locus))) %>%
  rowid_to_column("RowID")

# Add a column with a list of sample IDs where Reads > 0 for each locus
summary_data <- amplicons_negative %>%
  filter(Reads > 0) %>%
  group_by(Locus) %>%
  summarise(samples_with_reads = paste(unique(SampleID), collapse = ", "), .groups = "drop") %>%
  right_join(summary_data, by = "Locus")

# Reduce x-axis labels: label every nth locus
nth_label <- 20
summary_data <- summary_data %>%
  mutate(Locus_label = ifelse(RowID %% nth_label == 1, paste0(Chromosome, ":", Start), ""))

# Create the bar plot with hover text
p <- ggplot(summary_data, aes(x = Locus, y = sum_reads, fill = reaction,
                               text = paste("Locus: ", Locus, "<br>Total Reads: ", sum_reads,
                                          "<br>Controls with Reads: ", samples_with_reads))) +
  geom_bar(stat = "identity", position = "dodge") +
  xlab("Locus (Chromosome:Start Coordinate)") +
  ylab("Sum of Reads Across Controls") +
  ggtitle("Total Reads per Locus Across Negative Controls") +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 10),
    axis.title = element_text(size = 12),
    panel.grid = element_blank(),
    panel.background = element_blank(),
    legend.position = "top"
  ) +
  scale_x_discrete(labels = summary_data$Locus_label) +
  scale_fill_manual(values = reaction_colours)

# Convert to interactive plotly plot
ggplotly(p, tooltip = "text")
```

## Successul Amplification

### Parasitemia vs. Number of Successfully Amplified Loci

::: {.callout-tip collapse="true"}
## About this section
This plot examines the relationship between a sample's parasitemia (qPCR-determined parasite load) and the number of loci that successfully amplified. This plot helps assess whether lower parasitemia samples struggle with amplification, which can indicate potential limitations in sensitivity.
:::

```{r}
generate_parasitemia_by_amplification_success_plot <- function(summary_samples_for_batch, threshold, sample_colours, batch_name) {
    parasitemia_plot <- ggplot(data = summary_samples_for_batch) +
      # Plot points with different colors for each sample type
      geom_point(aes(x = Parasitemia + 0.9,
                     y = prop_good_loci,
                     color = SampleType),
                 shape = 1,
                 alpha=0.8,
                 stroke = 1) +
    # Log scale for the x-axis
    scale_x_log10() +
    ylim(0,1) +
    # Faceting by Batch column
    facet_grid(rows = vars(reaction),
               scale = "free_y")+
    # Adding labels and title
    ylab(paste0("Amplicons with >",threshold," reads")) +
    xlab("Parasitemia (log10)") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))+

    ggtitle(paste0(batch_name, " Amplicons with `good` read depth"))+
    # Color scale for SampleType
    scale_color_manual(values = sample_colours)
}

generate_parasitemia_by_amplification_success_plots <- function(summary_samples, threshold, sample_colours) {
    plots = list()
    for(b in unique(summary_samples$Batch)){
      summary_samples_batch = summary_samples %>%
        filter(Batch == b)
      plots[[b]] = generate_parasitemia_by_amplification_success_plot(summary_samples_batch, read_threshold, sample_colours, b)
    }
    return(plots)
}
```

```{r}
#| results: asis
#| echo: false
#| fig-column: screen-inset
plots <- generate_parasitemia_by_amplification_success_plots(summary_samples, read_threshold, sample_colours)
if (render_mode) {
  cat(create_tabsetOfGgplotObjects(plots))
} else {
  for (plot in plots) {
    print(plot)
  }
}
```

### Sample Reads vs. Number of Successfully Amplified Loci

::: {.callout-tip collapse="true"}
## About this section
This illustrates the relationship between the total number of reads per sample and the number of loci that passed the amplification threshold for that sample. This plot helps evaluate whether samples with higher read counts achieve better amplification success and can reveal potential issues such as insufficient sequencing depth or inefficient amplification. Ideally, a positive correlation should be observed, where higher read counts result in more successfully amplified loci.
:::

```{r}
generate_reads_by_amplification_success_plot <- function(summary_samples_for_batch, amplification_threshold,
                                                         sample_colours, reprep_threshold, repool_threshold, batch_name) {

    amp_plot <- ggplot(data = summary_samples_for_batch) +
      # Plot points with different colors for each sample type
      geom_point(aes(x = reads_per_sample + 0.9,
                     y=prop_good_loci,
                     color = SampleType),
                 shape=1,
                 alpha=0.8,
                 stroke = 1) +
    # Log scale for the x-axis
    scale_x_log10() +
    # Fix y-axis scale from 0 to 1
    scale_y_continuous(limits = c(0, 1)) +
    # Faceting by Batch column
    facet_grid(rows = vars(reaction),
               scale = "free_y")+
    # Adding labels and title
    ylab(paste0("Amplicons with >",amplification_threshold," reads")) +
    xlab("Total Reads for Sample") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))+

    ggtitle(paste0(batch_name, " Targets that Amplified Successfully"))+
    # Color scale for SampleType
    scale_color_manual(values = sample_colours) +
    # Add dashed horizontal lines at reprep_threshold and repool_threshold
    geom_hline(yintercept = reprep_threshold, linetype = "dashed") +
    geom_hline(yintercept = repool_threshold, linetype = "dashed") +
    # Add labels for the horizontal lines
    annotate("text", x = min(summary_samples_for_batch$reads_per_sample, na.rm = TRUE),
               y = reprep_threshold, label = "Reprep", vjust = 1, hjust = 0) +
    annotate("text", x = min(summary_samples_for_batch$reads_per_sample, na.rm = TRUE),
               y = repool_threshold, label = "Repool", vjust = 1, hjust = 0)
}

generate_reads_by_amplification_success_plots <- function(summary_samples, read_threshold, sample_colours, reprep_threshold, repool_threshold) {
    plots = list()
    for(b in unique(summary_samples$Batch)){
      summary_samples_batch = summary_samples %>%
        filter(Batch == b)
      plots[[b]] = generate_reads_by_amplification_success_plot(summary_samples_batch, read_threshold, sample_colours, reprep_threshold, repool_threshold, b)
    }
    return(plots)
}
```

```{r}
#| results: asis
#| echo: false
#| fig-column: screen-inset
read_plots <- generate_reads_by_amplification_success_plots(summary_samples, read_threshold, sample_colours, reprep_threshold, repool_threshold)
if (render_mode) {
  cat(create_tabsetOfGgplotObjects(read_plots))
} else {
  for (plot in read_plots) {
    print(plot)
  }
}
```

::: {.callout-note collapse="true"}
## Section D: Standardized Output Generation

```{r}
generate_reprep_repool_table <- function(summary_samples, reprep_threshold, repool_threshold) {
    summary_table <- summary_samples %>%
      mutate(
          status = case_when(
          prop_good_loci < reprep_threshold ~ "reprep",
          prop_good_loci < repool_threshold ~ "repool",
          TRUE ~ "pass"
        ),
        reason = case_when(
          prop_good_loci < reprep_threshold ~ paste("< reprep threshold", reprep_threshold),
          prop_good_loci < repool_threshold ~ paste("< repool threshold", repool_threshold),
          TRUE ~ "NA"
        )
      ) %>%
      arrange(desc(status)) %>%
      select(SampleID, Batch, reaction, status, reason, reads_per_reaction, prop_good_loci)

    return(summary_table)
}
```

```{r}
# Add in any samples that were in manifest and not in results
fill_missing_data <- function(reprep_repool_summary, manifest) {
  # Get unique reactions
  unique_reactions <- unique(reprep_repool_summary$reaction)

  # Identify missing samples from manifest
  missing_samples <- manifest %>%
    filter(!(SampleID %in% reprep_repool_summary$SampleID)) %>%
    select(SampleID, Batch)

  # Create rows for missing SampleIDs with all reactions
  missing_sample_rows <- missing_samples %>%
    crossing(reaction = unique_reactions) %>%
    mutate(
      status = "reprep",
      reason = "no data for sample",
      reads_per_reaction = 0,
      prop_good_loci = 0
    )

  # Ensure all existing samples have all reactions
  missing_reaction_rows <- reprep_repool_summary %>%
    select(SampleID, Batch) %>%
    distinct() %>%
    crossing(reaction = unique_reactions) %>%
    anti_join(reprep_repool_summary, by = c("SampleID", "Batch", "reaction")) %>%
    mutate(
      status = "reprep",
      reason = "no data for reaction",
      reads_per_reaction = 0,
      prop_good_loci = 0
    )

  # Combine everything
  bind_rows(reprep_repool_summary, missing_sample_rows, missing_reaction_rows)
}
```

```{r}
reprep_repool_summary <- generate_reprep_repool_table(summary_samples, reprep_threshold, repool_threshold)
reprep_repool_summary <- fill_missing_data(reprep_repool_summary, manifest)
head(reprep_repool_summary, n = 20) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r}
# Calculate pass percentage
repreps_count <- reprep_repool_summary %>%
  filter(status == "reprep", !grepl("Control", SampleID)) %>%
  distinct(SampleID) %>%
  nrow()

repools_count <- reprep_repool_summary %>%
  filter(status == "repool", !grepl("Control", SampleID)) %>%
  distinct(SampleID) %>%
  nrow()

passedQC_count <- reprep_repool_summary %>%
  filter(status == "pass", !grepl("Control", SampleID)) %>%
  distinct(SampleID) %>%
  nrow()

total_samples <- repreps_count + repools_count + passedQC_count
if (total_samples > 0) {
  percentage_pass <- (passedQC_count / total_samples) * 100
  cat("Overall QC Summary:\n")
  cat("  Reprep:", repreps_count, "\n")
  cat("  Repool:", repools_count, "\n")
  cat("  Pass:", passedQC_count, "\n")
  cat("  Pass percentage:", round(percentage_pass, 1), "%\n")
}
```
:::

## Amplification Plate Maps

### Reads Heatmap

::: {.callout-tip collapse="true"}
## About this section
The reads heatmap provides a visual representation of read distribution across the plate, helping to identify inconsistencies in sequencing efficiency. This can highlight potential issues such as edge effects, batch effects, or pipetting errors that may impact data quality and interpretation.
:::

```{r}
plot_plate_with_feature <- function(quadrants_batch, sample_colours, batch_name, fill_param, scale_midpoint, scale_label) {
    plate_layout <- ggplot(quadrants_batch) +
      # Heatmap fill based on reads_per_reaction
      geom_tile(aes(x = Column, y = y, fill = !!rlang::parse_expr(fill_param)), color = NA, width = 0.95, height = 0.95) +
      # Separate tile layer for outlines
      geom_rect(data = quadrants_batch,
            aes(xmin = Column-0.45, xmax = Column+0.45, ymin = y-0.45, ymax = y+0.45,color=SampleType),fill = NA,linewidth=1.3)+
      # Facet by Batch and reaction
      facet_grid(reaction ~ Batch) +
      # Color gradient for heatmap
      scale_fill_gradient2(
        low = "black",
        mid = "darkorange4",
        high = "darkorange",
        midpoint = scale_midpoint,
        name = scale_label
      ) +
      # Outline colors based on SampleType
      scale_color_manual(values = sample_colours, na.translate = TRUE) +
      # Labels
      scale_y_continuous(breaks = quadrants_batch$y, labels = quadrants_batch$Row) +
      scale_x_continuous(breaks = 1:12, labels = as.character(1:12)) +
      xlab("Column") +
      ylab("Row") +
      # Theme adjustments
      theme_minimal() +
      theme(
        axis.text = element_text(size = 12),
        strip.text = element_text(size = 14, face = "bold"),
        panel.grid = element_blank(),
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        legend.key = element_rect(colour = "black")
      ) +
    coord_fixed(ratio = 0.75)
    return(plate_layout)
}

plot_plate_heat_maps <- function(sample_data, quadrants, sample_colours, fill_param, scale_midpoint, scale_label) {
    plate_maps = list()

    quadrants_by_reaction <- crossing(summary_samples %>% select(reaction) %>% distinct() , quadrants)
    merged_data <- merge(summary_samples, quadrants_by_reaction, by = c("Batch", "Row", "Column", "SampleType", "reaction"), all.y = TRUE)

    # Ensure NAs in SampleType are labeled
    merged_data$SampleType[is.na(merged_data$SampleType)] <- "empty"

    # Remove rows with NA reactions (artifacts from the crossing/merge operation)
    merged_data <- merged_data %>% filter(!is.na(reaction))

    for(b in unique(merged_data$Batch)){
      merged_data_batch = merged_data %>%
        filter(Batch == b)
      plate_maps[[b]] = plot_plate_with_feature(merged_data_batch, sample_colours, b, fill_param, scale_midpoint, scale_label)
    }
    return(plate_maps)
}
```

```{r}
#| results: asis
#| echo: false
#| fig-column: screen-inset
plate_maps <- plot_plate_heat_maps(summary_samples, quadrants, sample_colours, "log10(reads_per_reaction + 0.1)", 2, "log10(reads_per_reaction)")
if (render_mode) {
  cat(create_tabsetOfGgplotObjects(plate_maps))
} else {
  for (plate in plate_maps) {
    print(plate)
  }
}
```

### Successful Amplification Heatmap

::: {.callout-tip collapse="true"}
## About this section
Here we visualise the success rate of amplification across the plate, allowing for the identification of poorly amplified regions or wells. This is useful for assessing the quality of the PCR process, ensuring that loci across all samples are adequately amplified, and helping to spot potential issues with specific wells or sample groups.
:::

```{r}
#| results: asis
#| echo: false
#| fig-column: screen-inset
plate_maps <- plot_plate_heat_maps(summary_samples, quadrants, sample_colours, "prop_good_loci", 0.5, "prop_good_loci")
if (render_mode) {
  cat(create_tabsetOfGgplotObjects(plate_maps))
} else {
  for (plate in plate_maps) {
    print(plate)
  }
}
```

## Save Output Files

::: {.callout-tip collapse="true"}
## About this section
All output files are generated consistently for both pipeline modes.
:::

```{r}
# Create output directories if they don't exist
output_dirs <- unique(dirname(c(
  reprep_repool_summary_path,
  polyclonal_information_path,
  neg_control_information_path,
  filtered_allele_data_path,
  missing_loci_report_path,
  samples_to_repool_path,
  samples_to_reprep_path,
  missing_samples_report_path
)))

for (dir in output_dirs) {
  if (!dir.exists(dir)) {
    dir.create(dir, recursive = TRUE)
    cat("Created output directory:", dir, "\n")
  }
}

# Write reprep/repool summary
write.csv(reprep_repool_summary, reprep_repool_summary_path, row.names = FALSE)
cat("Saved:", reprep_repool_summary_path, "\n")

# Write polyclonal controls
write.csv(polyclonal_information, polyclonal_information_path, row.names = FALSE)
cat("Saved:", polyclonal_information_path, "\n")

# Write negative control contamination
write.csv(neg_control_information, neg_control_information_path, row.names = FALSE)
cat("Saved:", neg_control_information_path, "\n")

# Create separate CSV files for samples needing repool and reprep
samples_to_repool <- reprep_repool_summary %>%
  filter(status == "repool") %>%
  select(SampleID, Batch, reaction, reads_per_reaction, prop_good_loci, reason) %>%
  arrange(Batch, SampleID, reaction)

samples_to_reprep <- reprep_repool_summary %>%
  filter(status == "reprep") %>%
  select(SampleID, Batch, reaction, reads_per_reaction, prop_good_loci, reason) %>%
  arrange(Batch, SampleID, reaction)

write.csv(samples_to_repool, samples_to_repool_path, row.names = FALSE)
cat("Saved:", samples_to_repool_path, "(", nrow(samples_to_repool), "samples needing repool)\n")

write.csv(samples_to_reprep, samples_to_reprep_path, row.names = FALSE)
cat("Saved:", samples_to_reprep_path, "(", nrow(samples_to_reprep), "samples needing reprep)\n")

# CRITICAL: Use allele_data_original (not allele_data_qc) to include excluded loci
# Read in raw data with no other filtering applied
allele_data_for_output <- read.csv(allele_data_path, sep='\t')

# Standardize SampleID
allele_data_for_output <- allele_data_for_output %>%
  mutate(SampleID = if (standardise_sampleID) SampleID %>%
           sub("_L001$", "", .) %>%
           word(., start = 1, end = -2, sep = "_") else SampleID)

# Identify samples in manifest but not in allele data
samples_in_allele_data <- allele_data_for_output %>%
  distinct(SampleID) %>%
  pull(SampleID)

missing_samples <- manifest %>%
  filter(!(SampleID %in% samples_in_allele_data)) %>%
  select(SampleID, Batch, SampleType, Row, Column, Parasitemia) %>%
  arrange(Batch, SampleID)

write.csv(missing_samples, missing_samples_report_path, row.names = FALSE)
cat("Saved:", missing_samples_report_path, "(", nrow(missing_samples), "samples in manifest not found in allele data)\n")

# If using pseudocigar column, then group by this for reads
if (allele_col == "PseudoCIGAR") {
  allele_data_for_output <- allele_data_for_output %>%
      group_by(SampleID, Locus, PseudoCIGAR) %>%
      summarise(Reads = sum(Reads), .groups = "drop")
}

# Calculate allele frequency
allele_data_for_output <- allele_data_for_output %>%
  group_by(SampleID, Locus) %>%
  mutate(AlleleFreq = Reads / sum(Reads)) %>%
  ungroup()

# Filter to minimum read and within sample allele frequency (if enabled)
if (filter_by_wsaf) {
  allele_data_filtered <- allele_data_for_output %>%
    filter(Reads > read_filter & AlleleFreq > af_filter)
  filter_msg <- paste0("Reads > ", read_filter, " AND WSAF > ", af_filter)
} else {
  # Only filter by read count, not WSAF
  allele_data_filtered <- allele_data_for_output %>%
    filter(Reads > read_filter)
  filter_msg <- paste0("Reads > ", read_filter, " (WSAF filter disabled)")
}

# Write out table
write.table(allele_data_filtered, filtered_allele_data_path, sep="\t", row.names = FALSE, quote = FALSE)
cat("Saved:", filtered_allele_data_path, "\n")
cat("  (Includes ALL loci, including those excluded from QC calculations)\n")
cat("  Filtered by:", filter_msg, "\n")
```

::: {.callout-note collapse="true"}
## Section F: Missing Loci Report (Conditional)

```{r}
# Missing loci report (attempt for both modes if pool_info exists)
if (!is.null(pool_info)) {
  loci_not_in_amplicon_coverage <- pool_info %>%
    anti_join(amplicon_coverage_qc, by = "Locus") %>%
    select(Locus, Pools, Category, chromosome)

  write.csv(loci_not_in_amplicon_coverage, missing_loci_report_path, row.names = FALSE)
  cat("Saved:", missing_loci_report_path, "\n")

  if (nrow(loci_not_in_amplicon_coverage) > 0) {
    cat("\n## Missing Loci Report\n\n")
    cat("The following loci are present in the pool file but missing from the amplicon coverage data:\n\n")

    missing_loci_table <- loci_not_in_amplicon_coverage %>%
      arrange(Locus) %>%
      mutate(No = row_number()) %>%
      select(No, everything()) %>%
      gt() %>%
      tab_header(
        title = "Loci Present in Pool File but Missing from Data"
      )
    print(missing_loci_table)
  } else {
    cat("\nAll loci in pool file were found in the data.\n")
  }
} else {
  cat("\nNo pool file available for missing loci report.\n")
}
```
:::

## QC Summary by Batch

::: {.callout-tip collapse="true"}
## About this section
This visualization shows the overall QC pass rate by batch, with stacked bars representing the percentage of samples passing, requiring repool, or requiring reprep. This helps identify batches with systematic issues.
:::

```{r}
#| fig-width: 10
#| fig-height: 6

# Calculate QC status counts by batch
qc_summary_by_batch <- reprep_repool_summary %>%
  filter(!grepl("Control", SampleID)) %>%  # Exclude controls
  distinct(SampleID, Batch, status) %>%
  group_by(Batch, status) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(Batch) %>%
  mutate(
    total = sum(count),
    percentage = (count / total) * 100
  ) %>%
  ungroup()

# Define colors for QC status
qc_colors <- c(
  "pass" = "darkgreen",
  "repool" = "orange",
  "reprep" = "red3"
)

# Create stacked bar plot
qc_plot <- ggplot(qc_summary_by_batch, aes(x = Batch, y = percentage, fill = status)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.3) +
  geom_text(
    aes(label = ifelse(percentage > 5, paste0(round(percentage, 1), "%"), "")),
    position = position_stack(vjust = 0.5),
    size = 3.5,
    fontface = "bold",
    color = "white"
  ) +
  scale_fill_manual(
    values = qc_colors,
    name = "QC Status",
    breaks = c("pass", "repool", "reprep"),
    labels = c("Pass", "Repool", "Reprep")
  ) +
  labs(
    title = "QC Status by Batch",
    x = "Batch",
    y = "Percentage of Samples (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.position = "top",
    panel.grid.major.x = element_blank()
  ) +
  ylim(0, 100)

print(qc_plot)
```

```{r}
# Print summary table
qc_summary_table <- reprep_repool_summary %>%
  filter(!grepl("Control", SampleID)) %>%
  distinct(SampleID, Batch, status) %>%
  group_by(Batch, status) %>%
  summarize(count = n(), .groups = "drop") %>%
  # Ensure all status levels exist for each batch
  complete(Batch, status = c("pass", "repool", "reprep"), fill = list(count = 0)) %>%
  pivot_wider(names_from = status, values_from = count, values_fill = 0) %>%
  mutate(
    total = pass + repool + reprep,
    pass_rate = round(100 * pass / total, 1)
  ) %>%
  select(Batch, pass, repool, reprep, total, pass_rate)

qc_summary_table %>%
  kable(col.names = c("Batch", "Pass", "Repool", "Reprep", "Total", "Pass Rate (%)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Summary

::: {.callout-important collapse="false"}
## Report Summary

This unified QC report has processed data in **`r pipeline_mode` mode** and generated the following standardized outputs:

1. **`r reprep_repool_summary_path`** - Combined reprep/repool classification for all samples
2. **`r samples_to_repool_path`** - Samples requiring repool
3. **`r samples_to_reprep_path`** - Samples requiring reprep
4. **`r polyclonal_information_path`** - Polyclonality information for positive controls
5. **`r neg_control_information_path`** - Contamination detected in negative controls
6. **`r filtered_allele_data_path`** - Filtered allele data meeting read/AF thresholds (includes all loci)
7. **`r missing_loci_report_path`** - Loci in pool file but absent from data (if applicable)
8. **`r missing_samples_report_path`** - Samples in manifest but not found in allele data

### Key Points

- **Loci filtered for QC**: `r length(excluded_loci)` loci were excluded from QC calculations
- **QC metrics**: Reprep/repool classifications based on `r n_distinct(amplicon_coverage_qc$Locus)` QC-relevant loci
- **Final output**: `allele_data_filtered.txt` includes ALL loci (both QC and excluded) meeting thresholds
- **Pass rate**: `r if(exists("percentage_pass")) paste0(round(percentage_pass, 1), "%") else "N/A"`
:::
