---
title: "QC Report"
author: ""
output:
  html_document:
    code-fold: true
    self-contained: true
    toc: true
execute:
  echo: false
---



This interactive notebook generates a summary of QC statistics to assess the success of a **Mad4hatter** run. 

::: {.callout-note collapse="true"}
## Required Inputs 

To proceed, you must provide the **results directory** from the Mad4hatter pipeline, which should include the following files:

* **sample_coverage.txt**
* **amplicon_coverage.txt**
* **allele_data.txt**
* **panel_information/amplicon_info.tsv**

Additionally, a **sample manifest** is required. This file must contain the following fields:

* **sample_name** – Unique identifier for each sample.
* **SampleType** – Specifies whether the entry is a **sample**, **positive** control, or **negative** control.
* **Batch** – Identifies a group of samples processed simultaneously by the same individual.
* **Column** – The well column where the sample was placed in the plate.
* **Row** – The well row where the sample was placed in the plate.
* **Parasitemia** – The qPCR value for the sample.
:::

::: {.callout-note collapse="true"}
## QC Summaries
The notebook provides the following analyses:

* **Plate Layout**: Location of samples and controls on the plate
* **Primer Dimer Content**: Input reads vs. % attributed to primer dimers 
* **Balancing Across Batches**: Swarm plot of reads output by the pipeline by batch
* **Control Summary**, including:  
    * Polyclonality of Positive Controls
    * Read Summary for Negative Controls
    * Contamination Maps for Negative Controls
* **Successul Amplification** summary, including:
    * Sample Reads vs. Number of Successfully Amplified Targets
    * Parasitemia vs. Number of Successfully Amplified Targets
* **Amplification Plate Maps**, including:
    * Reads Heatmap
    * Successful Amplification Heatmap
* **QC Summary by Batch**, including:
    * Proportion of samples for each status of QC (Pass, Repool, or Reprep) 
      
Note: A target is considered successfully amplified if it has more than `read_threshold` reads, where `read_threshold` is a threshold that can be set below.
:::


## Setup

::: {.callout-note collapse="true"}
### Configuration and Data Loading

```{r}
# Running mode - if running interactively, for example in Rstudio, set this to FALSE
render_mode <- TRUE
```

```{r}
# User input 
# Input path to results directory from Mad4hatter pipeline
results_dir <- "PATH/TO/MAD4HATTER_RESULTS_DIR/"
# Input path to manifest 
manifest_file <- "PATH/TO/MANIFEST"
output_dir <- ""
```

```{r}
# Path to outputs
reprep_repool_summary_path <- "reprep_repool_summary.csv"
polyclonal_information_path <- "positive_control_polyclonal_info.csv"
neg_control_information_path <- "negative_control_amplified_targets.csv"
filtered_allele_data_path <- "allele_data_filtered.txt"
filtered_collapsed_allele_data_path <- "allele_data_collapsed_filtered.txt"
missing_targets_report_path <- "missing_targets_report.csv"
samples_to_repool_path <- "samples_to_repool.csv"
samples_to_reprep_path <- "samples_to_reprep.csv"
missing_samples_report_path <- "missing_samples_report.csv"
filtered_resmarker_table_path <- "resmarker_table_filtered.txt"
filtered_resmarker_microhaplotype_table_path <- "resmarker_microhaplotype_table_filtered.txt"
```

```{r}
# If TRUE, trims extra sequencing-related information from SampleID to match the manifest format
standardise_sample_name <- TRUE
# Input a threshold for the reads needed to classify a target as having amplified successfully
read_threshold <- 100
# Input a minimum Reads per ASV used as a filter for the positive control plot
read_filter <- 0 
# Minimum Allele Frequency per ASV used as a filter for positive control plot 
af_filter <- 0.01 
# Read threshold to output target information for negative control
negative_control_read_threshold <- 50 
# Proportion of targets successfully amplified; samples below this threshold require re-prep
reprep_threshold <- 0.5 
# Proportion of targets successfully amplified; samples below this threshold require re-pool
repool_threshold <- 0.75
# This column indicates the one to be used for an allele ID, either asv or pseudocigar_masked. 
# If pseudocigar_masked, the table will be collapsed to unique pseudocigar_masked.
allele_col <- "pseudocigar_masked"
# Threshold length (including primers) to flag a target as being long. These targets will be filtered from performance metrics as they are known to have a lower performance
long_target_threshold <- 275
# If TRUE, af_filter will be applied to final output allele table. If FALSE, only read_filter will be applied 
filter_wsaf_final_allele_table <- TRUE
# If FALSE, the first digit of the pool will be used as the reaction. If TRUE, the pool will be used as the reaction 
reaction_as_pool <- FALSE
```

```{r}
# Print summary of user inputs
cat("============================================\n")
cat("USER CONFIGURATION SUMMARY\n")
cat("============================================\n")
cat("Results Directory:", results_dir, "\n")
cat("Manifest File:", manifest_file, "\n")
cat("Read Threshold for Successful Amplification:", read_threshold, "\n")
cat("Minimum Reads per ASV (Positive Control Filter):", read_filter, "\n")
cat("Minimum Allele Frequency per ASV (Positive Control Filter):", af_filter, "\n")
cat("Allele Column:", allele_col, "\n")
cat("============================================\n")
```

```{r}
# Create file paths 
sample_coverage_path <- file.path(results_dir,"sample_coverage.txt")
amplicon_coverage_path <- file.path(results_dir,"amplicon_coverage.txt")
allele_data_path <- file.path(results_dir, "allele_data.txt")
collapsed_allele_data_path <- file.path(results_dir, "allele_data_collapsed.txt")
panel_information_path <- file.path(results_dir, "panel_information/amplicon_info.tsv")
resmarker_table_path <- file.path(results_dir,"resistance_marker_module/resmarker_table.txt")
resmarker_microhaplotype_table_path <- file.path(results_dir,"resistance_marker_module/resmarker_microhaplotype_table.txt")
```

```{r}
# Load necessary libraries
library(ggplot2)
library(plotly, quietly = TRUE,warn.conflicts = FALSE)
library(dplyr, quietly = TRUE,warn.conflicts = FALSE)
library(tidyr)
library(ggbeeswarm)
library(stringr)
library(tibble)
library(purrr)
library(kableExtra, quietly = TRUE, warn.conflicts = FALSE)
```

```{r}
sep_guess <- if (any(grepl(";", readLines(allele_data_path, n = 1)))) ";" else "," # necessary for non-English format
manifest <- read.csv(manifest_file, sep = sep_guess) %>%
  mutate(Row = toupper(Row)) # Ensure Row is capitalised

sample_coverage <- read.csv(sample_coverage_path, sep="\t")
amplicon_coverage <- read.delim(amplicon_coverage_path)
allele_data <- read.csv(allele_data_path, sep='\t')
panel_information <- read.csv(panel_information_path, sep='\t')
```

```{r}
# Standardize sample names if requested
amplicon_coverage <- amplicon_coverage %>%
  mutate(sample_name = if (standardise_sample_name) sample_name %>%
           sub("_L001$", "", .) %>%
           word(., start = 1, end = -2, sep = "_") else sample_name)
allele_data <- allele_data %>%
  mutate(sample_name = if (standardise_sample_name) sample_name %>%
           sub("_L001$", "", .) %>%
           word(., start = 1, end = -2, sep = "_") else sample_name)
sample_coverage <- sample_coverage %>%
  mutate(sample_name = if (standardise_sample_name) sample_name %>%
           sub("_L001$", "", .) %>%
           word(., start = 1, end = -2, sep = "_") else sample_name)
```

:::

::: {.callout-note collapse="true"}
### Section A: Locus Filtering

Here we filter out two types of targets for QC calculations only:

* **Targets amplifying non-Plasmodium falciparum species**, which are expected to amplify only if other species are present.

* **Long amplicons (>long_target_threshold bp, including primers)**, which tend to underperform due to their length.

These targets should not be considered when assessing the success of a sequencing run, but will still be included in the final allele_data output.

```{r}
# Find long targets to remove 
panel_information <- panel_information %>%
  mutate(
    target_length = (insert_end - insert_start) +
                    nchar(fwd_primer) +
                    nchar(rev_primer)
  )

# Long loci 
long_loci <- panel_information %>%
  filter(target_length > long_target_threshold) %>%
  pull(target_name)

cat("Long targets being removed:\n")
cat(unlist(long_loci), sep = "\n")
```

```{r}
# Species Loci to remove 
old_species_loci_names <- c(
    'Pf3D7_13_v3-1041593-1041860-1AB',
    'PmUG01_12_v1-1397996-1398245-1AB',
    'PocGH01_12_v1-1106456-1106697-1AB',
    'PvP01_12_v1-1184983-1185208-1AB',
    'PKNH_12_v2-198869-199113-1AB'
)
new_species_loci_names <- c(
    'Pf3D7_13_v3-1041624-1041829',
    'PmUG01_12_v1-1398020-1398213',
    'PocGH01_12_v1-1106482-1106671',
    'PvP01_12_v1-1185007-1185184',
    'PKNH_12_v2-0198893-0199083'
)
M1addon_loci <- c(
    'PKNH_MIT_v2-0002298-0002488',
    'PmUG01_MIT_v1-0000606-0000784',
    'PocGH01_MIT_v2-0000358-0000550',
    'PvP01_MIT_v2-0004470-0004642'
)
```

```{r}
# Combine into a list of all unique excluded loci
excluded_loci <- c(old_species_loci_names, new_species_loci_names, M1addon_loci, long_loci)
excluded_loci <- unique(excluded_loci)

# Filter amplicon_coverage to exclude rows with loci in the exclusion list
amplicon_coverage_qc <- amplicon_coverage[!amplicon_coverage$target_name %in% excluded_loci, ]

# Filter allele_data to exclude rows with loci in the exclusion list
allele_data_qc <- allele_data[!allele_data$target_name %in% excluded_loci, ]

# Filter panel_information to exclude rows with loci in the exclusion list
panel_information_qc <- panel_information[!panel_information$target_name %in% excluded_loci, ]
```

```{r}
cat("\nTarget count summary:\n")
cat("  Total targets in amplicon_coverage:", n_distinct(amplicon_coverage$target_name), "\n")
cat("  Targets used for QC (after filtering):", n_distinct(amplicon_coverage_qc$target_name), "\n")
cat("  Excluded targets will still appear in final allele_data_filtered.txt output\n")
```

:::

::: {.callout-note collapse="true"}
### Section B: Pool File Loading & Reaction Assignment

```{r}
# Add reaction to amplicon table 
# reaction here refers to 1 of the 2 mPCR reactions. 
if (reaction_as_pool) {
  panel_information_qc <- panel_information_qc %>%
  mutate(
    reaction = pool
  )
} else {
  panel_information_qc <- panel_information_qc %>%
  mutate(
    reaction = (str_extract(pool, "\\d+"))
  )
}
```

```{r}
# Count loci per reaction (only including loci that were not filtered above)
nloci_table <- panel_information_qc %>% 
    count(reaction) %>% 
    rename(nreactionloci = n)
nloci_table |>
      kable() |>
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r}
# If you want to manually edit the target counts instead of using the automated draw from the results folder, then uncomment and edit this code
# reaction_names <- c()
# target_count <- c()

# # Put into a table
# nloci_table <- data.frame(
#   reaction = reaction_names,
#   nreactionloci = c(target_count)
# )
# nloci_table
```

:::

```{r}
merge_data <- function(manifest, sample_coverage, standardise_sample_name) {
    merged_data <- sample_coverage %>%
        mutate(reads = suppressWarnings(as.numeric(reads))) %>% 
        replace_na(list(reads = 0)) %>%
        pivot_wider(names_from = stage, values_from = reads) %>%
        inner_join(manifest, by = "sample_name") 
    return(merged_data)
}
```

```{r}
# Reformat sample_coverage and add sample information from manifest 
sample_coverage_with_manifest <- merge_data(manifest, sample_coverage, standardise_sample_name)
```

```{r}
# Add manifest to amplicon coverage
amplicon_coverage_with_manifest <- amplicon_coverage_qc %>%
    inner_join(manifest,by = c("sample_name")) %>% 
    left_join(panel_information_qc %>% select(target_name, reaction), by = "target_name") 
```

```{r}
# Create sample summary stats 
summary_samples <- amplicon_coverage_with_manifest %>%
  # Join the nloci_table
  left_join(nloci_table, by = "reaction") %>%
  # Calculate number of loci with reads over threshold per reaction
  group_by(sample_name, Batch, SampleType, reaction, nreactionloci) %>%
  summarize(
    reads_per_reaction = sum(OutputPostprocessing),
    n_good_loci = sum(OutputPostprocessing > read_threshold),
    .groups = "drop"
  ) %>%
    group_by(sample_name, Batch, SampleType) %>%
      mutate(
        reads_per_sample = sum(reads_per_reaction),
      ) %>%
  ungroup() %>%
  mutate(prop_good_loci = n_good_loci / nreactionloci) %>%
  inner_join(manifest, by = join_by(sample_name, Batch, SampleType))
```

```{r}
# Plotting Setup
sample_colours <- c(
  "negative" = "red3",
  "positive" = "blue3",
  "sample" = "darkgrey",
  "empty" = "black"
)
```

```{r}
reaction_colours <- c("orangered2","skyblue","black","red","green3","blue","cyan","magenta","yellow","gray")
```

```{r}
# Print summary of samples in files
cat("Sample Summary\n")
cat(length(unique(manifest$sample_name)), "samples in manifest", "\n")
cat(length(unique(allele_data$sample_name)), "samples in allele data file", "\n")
cat(length(unique(allele_data_qc$sample_name)), "samples in filtered allele data file", "\n")
cat(length(unique(sample_coverage$sample_name)), "samples in manifest and sample coverage file", "\n")
cat(length(unique(amplicon_coverage_qc$sample_name)), "samples in manifest and amplicon coverage file", "\n")
```

## Plate Layout

::: {.callout-tip collapse="true"}
## About this section
The plate layout provides a visual representation of sample organization, including the placement of positive and negative controls. This is allows verification that controls are positioned correctly to validate assay performance with later plots.
:::

```{r}
create_tabsetOfGgplotObjects <- function(ggplotObjectsList) {
  cat("\n\n::: {.panel-tabset}\n\n")
  iwalk(ggplotObjectsList, ~ {
    cat('## ', .y, '\n\n')
    print(.x)
    cat('\n\n')
  })
  cat(":::\n")
}
```

```{r}
create_plate_template <- function(summary_samples, batches, nrows=8, ncols=12) {
    quadrants <-
      expand.grid(
        Batch = batches,
        y = 1:nrows,
        Column = 1:ncols
      ) %>%
      mutate(
        ymin = y - 0.45,
        ymax = y + 0.45,
        xmin = Column - 0.45,
        xmax = Column + 0.45
      ) %>%
      mutate(Row = toupper(rev(letters[1:nrows])[y])) %>%
      # Join on the sample information to the plate
      left_join(
        summary_samples %>%
          select(Batch, Column, Row, SampleType) %>%
          distinct(),
        by = c("Batch", "Column", "Row" = "Row")
      )
    return(quadrants)
}
```

```{r}
quadrants <- create_plate_template(summary_samples, unique(summary_samples$Batch))
```

```{r}
plot_plate <- function(quadrants_batch, sample_colours, batch_name) {
    plate_layout <- ggplot(quadrants_batch) +
          # Tiles representing sample types, fill based on SampleType
          geom_tile(aes(x = Column, y = y, fill = SampleType), color = "white") +
          scale_y_continuous(breaks = quadrants_batch$y, labels = quadrants_batch$Row) +
          # Make sure labels are correct
          xlab("Column") +
          ylab("Row") +
          # Customizing the x-axis
          scale_x_continuous(breaks = 1:12, labels = as.character(1:12)) +
          # Quadrants as rectangles with borders
          geom_rect(
            data = quadrants_batch,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            color = "black", fill = NA, linewidth = 1
          ) +
          # Custom fill colors for each SampleType
          scale_fill_manual(
            values = sample_colours
          ) +
          # Theming
          theme_minimal() +
          theme(
            axis.text = element_text(size = 12),
            strip.text = element_text(size = 14, face = "bold"),
            panel.grid = element_blank(),  
            plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
          ) +
        # Customizing the legend to add an outline for the colored squares
        guides(
          fill = guide_legend(
            override.aes = list(
              color = "black",  # Outline color for the squares
              size = 1          # Outline size for the squares
            )
          )
        ) +
        # Title for the plot
        ggtitle(paste0("Plate Map: ", batch_name)) +
        coord_fixed(ratio = 0.75)
    return(plate_layout)
}

plot_plate_layouts <- function(plate_data, sample_colours) {
    plate_maps = list()
    plate_data$SampleType[is.na(plate_data$SampleType)] <- "empty"
    for(b in unique(plate_data$Batch)){
      quadrants_batch = plate_data %>%
        filter(Batch == b)
      plate_maps[[b]] = plot_plate(quadrants_batch, sample_colours, b)
    }
    return(plate_maps)
}
```

```{r}
#| results: asis
#| echo: false
#| fig-column: screen-inset
plate_maps <- plot_plate_layouts(quadrants, sample_colours)
if (render_mode) {
  cat(create_tabsetOfGgplotObjects(plate_maps))
} else {
  for (plate in plate_maps) {
    print(plate)
  }
}
```

## Primer Dimer Content

::: {.callout-tip collapse="true"}
## About this section
Here we visualise the proportion of sequencing reads that are classified as primer dimers, which occur when primers anneal to each other instead of the target DNA. This plot is useful for assessing the efficiency of the amplification process, as high primer dimer levels can indicate suboptimal reaction conditions, reduced sequencing efficiency, and potential issues with sample quality or reagent performance.
:::

```{r}
# Function to generate the plot
generate_dimer_plot <- function(sample_coverage_with_manifest, sample_colours) {
    
    dimer_plot <- ggplot(data = sample_coverage_with_manifest %>%  arrange(desc(SampleType))) +
    # Plot points with different colors for each sample type
      geom_point(aes(x = Input + 0.9,
                     y = (1 - OutputPostprocessing / Input) * 100,
                     color = SampleType),
                 shape = 1,
                 alpha=0.8,
                 stroke = 1) +
    # Log scale for the x-axis
    scale_x_log10() +
    # Faceting by Batch column
    facet_wrap(~Batch) +
    # Adding labels and title
    ylab("% Dimers") +
    xlab("Input Reads") +
    ggtitle("Dimer Content")+
    # Color scale for SampleType
    scale_color_manual(values = sample_colours)
    print(dimer_plot)
}
```

```{r}
generate_dimer_plot(sample_coverage_with_manifest, sample_colours)
```

## Balancing Across Batches

::: {.callout-tip collapse="true"}
## About this section
Here we show the distribution of total reads per sample across different batches, helping to assess whether sequencing depth is consistent. This plot is useful when you have multiple batches in the same sequencing run to identify imbalances in sequencing, which can arise due to variations in library preparation, loading efficiency, or sequencing conditions.
:::

```{r}
generate_balancing_plot <- function(summary_samples, sample_colours) {
  ggplot() +
    ggbeeswarm::geom_quasirandom(
    data = summary_samples %>%
        select(sample_name, Batch, reads_per_sample, SampleType) %>%
        distinct(),
    aes(x = Batch, y = reads_per_sample + 0.9, color = SampleType)
    ) +
    scale_y_log10() +
    scale_color_manual(values = sample_colours) +
    ylab("Total Reads for Sample") +
    ggtitle("Balancing Across Batches") +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1)
    )
}
```

```{r}
generate_balancing_plot(summary_samples, sample_colours)
```

## Control Summary

### Multiallelic Targets in Positive Controls

::: {.callout-tip collapse="true"}
## About this section
Here we inspect the positive controls, ensuring that the positive controls perform as expected without contamination or unwanted diversity. For example, if the control you included was monoclonal you would expect little to no targets reported as having more than one allele (multiallelic). In some cases you may allow for some level of false positive detection within a monoclonal control. Note that WSAF and total reads per allele filters (as set by default or modified by user) are applied.
:::

```{r}
# Add on within sample allele freqeucny 
allele_data_qc <- allele_data_qc %>%
  group_by(sample_name, target_name) %>%
  mutate(AlleleFreq = reads / sum(reads)) %>%
  ungroup()
```

```{r}
pos_control_allele_data <- allele_data_qc %>%
  inner_join(manifest %>% filter(SampleType == "positive"), by = "sample_name")
```

```{r}
if (allele_col=="pseudocigar_masked") {
  pos_control_data <- pos_control_allele_data %>%
      group_by(sample_name, target_name, pseudocigar_masked) %>%
      summarise(reads = sum(reads), AlleleFreq = sum(AlleleFreq), .groups = "drop")
} else {
  pos_control_data <- pos_control_allele_data
}
```

```{r}
# Apply filters and identify the number of alleles being called per locus
locus_summary <- pos_control_data %>%
  group_by(sample_name, target_name) %>%
  summarise(
    NumASVs_Meeting_Threshold = sum(reads > read_filter & AlleleFreq > af_filter),
    .groups = "drop"
  )

# Unified 3-category polyclonality classification (both modes)
locus_summary <- locus_summary %>%
  mutate(
    Category = case_when(
      NumASVs_Meeting_Threshold == 1 ~ "Monoallelic",
      NumASVs_Meeting_Threshold == 2 ~ "2 Alleles",
      NumASVs_Meeting_Threshold > 2  ~ ">2 Alleles"
    )
  )
fill_colors <- c("Monoallelic" = "Gray", "2 Alleles" = "orangered2", ">2 Alleles" = "cyan")

# Count loci per sample for each category
result <- locus_summary %>%
  group_by(sample_name, Category) %>%
  summarise(Count = n(), .groups = "drop")

# Create ggplot
p <- ggplot(result, aes(x = sample_name, y = Count, fill = Category, text = paste("Count:", Count))) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Alleles per Target for Positive Controls", x = "Sample Name", y = "Number of Targets") +
  theme_minimal() +
  scale_fill_manual(values = fill_colors) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)
  )

# Convert ggplot to interactive plot
ggplotly(p, tooltip = "text", width = 800, height = 500)
```

```{r}
# Filter data to keep only loci with more than one allele (polyclonal cases)
polyclonal_information <- pos_control_data %>%
  inner_join(locus_summary, by = c("sample_name", "target_name")) %>%
  filter(Category %in% c("2 Alleles", ">2 Alleles"))  # Keep only polyclonal categories

if (nrow(polyclonal_information) == 0) {
  cat("No multiallelic targets found for positive controls.\n")
} else {
    # Display as styled table
    head(polyclonal_information) |>
      kable() |>
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
}
```

### Negative Control Contamination 

#### Read Summary per Negative Control
::: {.callout-tip collapse="true"}
## About this section
In an ideal scenario, negative controls should have minimal or no amplification. If a negative control shows a high number of targets amplified with significant reads, it suggests potential contamination. By plotting the number of reads against the number of targets for each sample, any outliers or unexpected amplification in negative controls can be easily flagged.
:::

```{r}
# retrieve negative controls from amplicons.full
amplicons_negative = amplicon_coverage_with_manifest %>% 
  filter(SampleType=="negative") %>% 
  mutate(negative = paste0(Batch," Well: ",toupper(Row),Column))
```

```{r}
# Get negative controls grouped by batch
batches_with_negatives <- amplicons_negative %>%
  group_by(Batch) %>%
  summarize(
    n_controls = n_distinct(negative),
    .groups = "drop"
  ) %>%
  arrange(Batch)

cat("Negative controls by batch:\n")

batches_with_negatives |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

n_batches <- nrow(batches_with_negatives)

# Define layout parameters
ncol_layout <- 3         # Number of columns in facet grid
base_width <- 12
base_height_per_row <- 2.5  # Height per row of facets
min_height <- 6
max_height <- 20  # Prevent excessively tall figures

cat("Total batches with negative controls:", n_batches, "\n")

batches_with_negatives_with_reads <- amplicons_negative %>% 
filter(OutputPostprocessing!=0) %>%
  group_by(Batch) %>%
  summarize(
    n_controls = n_distinct(negative),
    .groups = "drop"
  ) %>%
  arrange(Batch)

n_batches_with_reads <- nrow(batches_with_negatives_with_reads)
cat("Total batches with negative controls that had reads:", n_batches, "\n")
```

```{r}
# Plotting helper function for negative control histograms
plot_negative_control_histogram <- function(data, reaction_colours, batch_name = NULL, remove_zeros = TRUE) {
  # ensure reaction is a factor
  data <- data %>% mutate(reaction = factor(reaction))

  # optionally remove zero-read targets
  if (remove_zeros) {
    data <- data %>% filter(!is.na(OutputPostprocessing) & OutputPostprocessing > 0)
  }

  # If nothing to plot, return NULL so caller can skip printing
  if (nrow(data) == 0) {
    return(NULL)
  }

  # compute x and y limits according to your rules
  max_read <- suppressWarnings(max(data$OutputPostprocessing, na.rm = TRUE))
  if (!is.finite(max_read)) max_read <- 0

  x_max <- if (max_read < 30) 40 else (max_read + 10)
  x_limits <- c(0, x_max)

  total_targets <- nrow(data)
  y_max <- if (total_targets < 90) 10 else (total_targets + 10)
  y_limits <- c(0.01, y_max)

  title_suffix <- if (!is.null(batch_name)) paste0(" - ", batch_name) else ""

  p <- ggplot(data) +
    geom_histogram(aes(x = OutputPostprocessing, fill = reaction), bins = 12, boundary = 0) +  # fewer bins
    facet_wrap(~ negative, ncol = ncol_layout, scales = "fixed") +
    xlab("Reads") +
    ylab("Number of targets") +
    ggtitle(paste0("Read Distribution per Negative Control", title_suffix)) +
    scale_fill_manual(values = reaction_colours) +
    coord_cartesian(xlim = x_limits, ylim = y_limits, expand = FALSE) +
    theme_minimal(base_size = 14) +
    theme(
      axis.text = element_text(size = 12),
      strip.text = element_text(size = 11, face = "bold"),
      panel.spacing = unit(1, "lines")
    )

  return(p)
}
```

```{r}
#| results: asis
#| echo: false

if (n_batches == 1) {
  # Single batch - no tabs needed, just dynamic height
  batch_data <- amplicons_negative

  total_negatives <- n_distinct(batch_data$negative)
  # If you only want to plot when there are reads, filter here first
  batch_data_nonzero <- batch_data %>% filter(!is.na(OutputPostprocessing) & OutputPostprocessing > 0)
  plotted_negatives <- n_distinct(batch_data_nonzero$negative)
  zero_read_negatives <- total_negatives - plotted_negatives

  if (nrow(batch_data_nonzero) == 0) {
    cat("No negative controls with reads in this batch — nothing to plot.\n")
  } else {
    # concise status line
    cat(sprintf("Plotting %d/%d negatives (%d zero-read)\n\n", plotted_negatives, total_negatives, zero_read_negatives))

    n_controls <- plotted_negatives

    # Calculate height for this batch
    batch_height <- min(max_height, max(min_height,
                                       ceiling(n_controls / ncol_layout) * base_height_per_row))

    knitr::opts_chunk$set(fig.width = base_width, fig.height = batch_height)

    p <- plot_negative_control_histogram(batch_data_nonzero, reaction_colours)
    if (!is.null(p)) print(p)
  }

} else {
  # Multiple batches - create tabs for each batch
  cat("\n\n::: {.panel-tabset}\n\n")

  batch_list <- unique(amplicons_negative$Batch)

  for (batch in batch_list) {
    cat('## ', batch, '\n\n')

    # Filter data for this batch
    batch_data <- amplicons_negative %>%
      filter(Batch == batch)

    total_negatives <- n_distinct(batch_data$negative)

    # Remove zero-read rows (we'll only plot batches with reads)
    batch_data_nonzero <- batch_data %>%
      filter(!is.na(OutputPostprocessing) & OutputPostprocessing > 0)

    plotted_negatives <- n_distinct(batch_data_nonzero$negative)
    zero_read_negatives <- total_negatives - plotted_negatives

    # If no reads in this batch, report and skip
    if (nrow(batch_data_nonzero) == 0) {
      cat("Skipping batch:\n", batch, "No reads present in any negative controls in this batch.\n\n")
      next
    }

    # concise status line
    cat(sprintf("Plotting reads for %d/%d negative controls (%d had no reads)\n\n", plotted_negatives, total_negatives, zero_read_negatives))

    # Calculate number of controls in this batch (only those with reads)
    n_controls_in_batch <- plotted_negatives

    # Calculate height for this batch
    batch_height <- min(max_height, max(min_height,
                                       ceiling(n_controls_in_batch / ncol_layout) * base_height_per_row))

    # Set figure dimensions for this batch
    knitr::opts_chunk$set(fig.width = base_width, fig.height = batch_height)

    # Create and print plot
    p <- plot_negative_control_histogram(batch_data_nonzero, reaction_colours, batch)
    if (!is.null(p)) print(p)

    cat('\n\n')
  }

  cat(":::\n\n")
}

# Reset chunk options to defaults
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
```

```{r}
cat("Summary of negative control targets with reads above threshold (",negative_control_read_threshold,") \n")
neg_control_information <- amplicons_negative %>%
    filter(OutputPostprocessing > negative_control_read_threshold) %>%
    select('sample_name','target_name','OutputPostprocessing')

if (nrow(neg_control_information) == 0) {
  cat("No negative controls had targets with reads over threshold.\n")
}
head(neg_control_information) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

#### Aggregated Reads per Target

::: {.callout-tip collapse="true"}
## About this section
This plot aggregates the total reads for each target across all negative controls. In negative controls, if there is an unexpected spike in reads for specific targets, it could indicate contamination in the form of cross-sample contamination or environmental contamination. Analyzing these total summed reads helps pinpoint specific targets where contamination might have occurred, offering insights into which steps in the process might have introduced contaminants.
:::

```{r}
# Get reads for each locus, label reaction and extract chrom and start from locus name
summary_data <- amplicons_negative %>%
  group_by(reaction, target_name) %>%
  summarise(sum_reads = sum(reads, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    Chromosome = sub("-.*", "", target_name),
    Start = as.numeric(sub(".*-(\\d+)-.*", "\\1", target_name))
  ) %>%
  arrange(Chromosome, Start) %>%
  mutate(target_name = factor(target_name, levels = unique(target_name))) %>%
  mutate(reaction = factor(reaction)) %>%
  rowid_to_column("RowID")

# Add a column with a list of sample IDs where Reads > 0 for each locus
summary_data <- amplicons_negative %>%
  filter(reads > 0) %>%
  group_by(target_name) %>%
  summarise(samples_with_reads = paste(unique(sample_name), collapse = ", "), .groups = "drop") %>%
  right_join(summary_data, by = "target_name")

# Reduce x-axis labels: label every nth locus
nth_label <- 20
summary_data <- summary_data %>%
  mutate(Locus_label = ifelse(RowID %% nth_label == 1, paste0(Chromosome, ":", Start), ""))

# Create the bar plot with hover text
p <- ggplot(summary_data, aes(x = target_name, y = sum_reads, fill = reaction, text = paste("Target: ", target_name, "<br>Total Reads: ", sum_reads, "<br>Controls with Reads: ", samples_with_reads))) +
  geom_bar(stat = "identity", position = "dodge") +
  xlab("Target (Chromosome:Start Coordinate)") +
  ylab("Sum of Reads Across Controls") +
  ggtitle("Total Reads per Target Across Negative Controls") +
theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 8),
    axis.title = element_text(size = 12),
    panel.grid = element_blank(),
    panel.background = element_blank(),
    legend.position = "right"
  ) +
  scale_x_discrete(labels = summary_data$Locus_label) +  # Apply reduced labels
  scale_fill_manual(values = reaction_colours)

# Convert to interactive plotly plot
ggplotly(p, tooltip = "text")
```

## Successul Amplification

### Parasitemia vs. Number of Successfully Amplified Targets
::: {.callout-tip collapse="true"}
## About this section
This plot examines the relationship between a sample's parasitemia (qPCR-determined parasite load) and the number of targets that successfully amplified. This plot helps assess whether lower parasitemia samples struggle with amplification, which can indicate potential limitations in sensitivity.
:::

```{r}
generate_parasitemia_by_amplification_success_plot <- function(summary_samples_for_batch, threshold, sample_colours, batch_name) {
    parasitemia_plot <- ggplot(data = summary_samples_for_batch) +
      # Plot points with different colors for each sample type
      geom_point(aes(x = Parasitemia + 0.9, 
                     y = prop_good_loci, 
                     color = SampleType),
                 shape = 1,
                 alpha=0.8,
                 stroke = 1) +
    # Log scale for the x-axis
    scale_x_log10() +
    ylim(0,1) + 
    # Faceting by Batch column
    facet_grid(rows = vars(reaction),
               scale = "free_y")+
    # Adding labels and title
    ylab(paste0("Amplicons with >",threshold," reads")) +
    xlab("Parasitemia (log10)") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))+

    ggtitle(paste0(batch_name, " Amplicons with `good` read depth"))+
    # Color scale for SampleType
    scale_color_manual(values = sample_colours) 
}

generate_parasitemia_by_amplification_success_plots <- function(summary_samples, threshold, sample_colours) {
    plots = list()
    for(b in unique(summary_samples$Batch)){
      summary_samples_batch = summary_samples %>%
        filter(Batch == b)# %>%
      plots[[b]] = generate_parasitemia_by_amplification_success_plot(summary_samples_batch, read_threshold, sample_colours, b)
    }
    return(plots)
}
```

```{r}
#| results: asis
#| echo: false
#| fig-column: screen-inset
plots <- generate_parasitemia_by_amplification_success_plots(summary_samples, read_threshold, sample_colours)
if (render_mode) {
  cat(create_tabsetOfGgplotObjects(plots))
} else {
  for (plot in plots) {
    print(plot)
  }
}
```

### Sample Reads vs. Number of Successfully Amplified Targets

::: {.callout-tip collapse="true"}
## About this section
This illustrates the relationship between the total number of reads per sample and the number of targets that passed the amplification threshold for that sample. This plot helps evaluate whether samples with higher read counts achieve better amplification success and can reveal potential issues such as insufficient sequencing depth or inefficient amplification. Ideally, a positive correlation should be observed, where higher read counts result in more successfully amplified targets.
:::

```{r}
generate_reads_by_amplification_success_plot <- function(summary_samples_for_batch, amplification_threshold, 
                                                         sample_colours, reprep_threshold, repool_threshold, batch_name) {
    
    amp_plot <- ggplot(data = summary_samples_for_batch) +
      # Plot points with different colors for each sample type
      geom_point(aes(x = reads_per_sample + 0.9,  
                     y=prop_good_loci,
                     color = SampleType),
                 shape=1,
                 alpha=0.8,
                 stroke = 1) +
    # Log scale for the x-axis
    scale_x_log10() +
    # Fix y-axis scale from 0 to 1
    scale_y_continuous(limits = c(0, 1)) +
    # Faceting by Batch column
    facet_grid(rows = vars(reaction),
               scale = "free_y")+
    # Adding labels and title
    ylab(paste0("Amplicons with >",amplification_threshold," reads")) +
    xlab("Total Reads for Sample") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))+

    ggtitle(paste0(batch_name, " Targets that Amplified Successfully"))+
    # Color scale for SampleType
    scale_color_manual(values = sample_colours) +
    # Add dashed horizontal lines at reprep_threshold and repool_threshold
    geom_hline(yintercept = reprep_threshold, linetype = "dashed") +
    geom_hline(yintercept = repool_threshold, linetype = "dashed") +
    # Add labels for the horizontal lines
    annotate("text", x = min(summary_samples_for_batch$reads_per_sample, na.rm = TRUE), 
               y = reprep_threshold, label = "Reprep", vjust = 1, hjust = 0) +
    annotate("text", x = min(summary_samples_for_batch$reads_per_sample, na.rm = TRUE), 
               y = repool_threshold, label = "Repool", vjust = 1, hjust = 0) 
}

generate_reads_by_amplification_success_plots <- function(summary_samples, read_threshold, sample_colours, reprep_threshold, repool_threshold) {
    plots = list()
    for(b in unique(summary_samples$Batch)){
      summary_samples_batch = summary_samples %>%
        filter(Batch == b)
      plots[[b]] = generate_reads_by_amplification_success_plot(summary_samples_batch, read_threshold, sample_colours, reprep_threshold, repool_threshold, b)
    }
    return(plots)
}
```

```{r}
#| results: asis
#| echo: false
#| fig-column: screen-inset
read_plots <- generate_reads_by_amplification_success_plots(summary_samples, read_threshold, sample_colours, reprep_threshold, repool_threshold)
if (render_mode) {
  cat(create_tabsetOfGgplotObjects(read_plots))
} else {
  for (plot in read_plots) {
    print(plot)
  }
}
```

::: {.callout-note collapse="true"}
## Per Sample QC Outcome Summary

```{r}
generate_reprep_repool_table <- function(summary_samples, reprep_threshold, repool_threshold) {
    summary_table <- summary_samples %>%
      filter(SampleType!='negative') %>%
      mutate(
          status = case_when(
          prop_good_loci < reprep_threshold ~ "reprep",
          prop_good_loci < repool_threshold ~ "repool",
          TRUE ~ "pass"
        ),
        reason = case_when(
          prop_good_loci < reprep_threshold ~ paste("< reprep threshold ", reprep_threshold),
          prop_good_loci < repool_threshold ~ paste("< repool threshold ", repool_threshold),
          TRUE ~ "NA"
        )
      ) %>%
      arrange(desc(status)) %>%
      select(sample_name, Batch, reaction, status, reason, reads_per_reaction, prop_good_loci)
    
    return(summary_table)
}
```

```{r}
# Add in any samples that were in manifest and not in results
fill_missing_data <- function(reprep_repool_summary, manifest) {
  # Get unique reactions
  unique_reactions <- unique(reprep_repool_summary$reaction)

  # Identify missing samples from manifest
  missing_samples <- manifest %>%
    filter(!(sample_name %in% reprep_repool_summary$sample_name)) %>%
    select(sample_name, Batch)

  # Create rows for missing SampleIDs with all reactions
  missing_sample_rows <- missing_samples %>%
    crossing(reaction = unique_reactions) %>%
    mutate(
      status = "reprep",
      reason = "no data for sample",
      reads_per_reaction = 0,
      prop_good_loci = 0
    )
  
  # Ensure all existing samples have all reactions
  missing_reaction_rows <- reprep_repool_summary %>%
    select(sample_name, Batch) %>%
    distinct() %>%
    crossing(reaction = unique_reactions) %>%
    anti_join(reprep_repool_summary, by = c("sample_name", "Batch", "reaction")) %>%
    mutate(
      status = "reprep",
      reason = "no data for reaction",
      reads_per_reaction = 0,
      prop_good_loci = 0
    )

  # Combine everything
  bind_rows(reprep_repool_summary, missing_sample_rows, missing_reaction_rows)
}
```

```{r}
#| scrolled: true
reprep_repool_summary <- generate_reprep_repool_table(summary_samples, reprep_threshold, repool_threshold)
reprep_repool_summary <- fill_missing_data(reprep_repool_summary, manifest)
reprep_repool_summary %>% filter(status!='pass') |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r}
# Calculate pass percentage
repreps_count <- reprep_repool_summary %>%
  filter(status == "reprep", !grepl("Control", sample_name)) %>%
  distinct(sample_name) %>%
  nrow()

repools_count <- reprep_repool_summary %>%
  filter(status == "repool", !grepl("Control", sample_name)) %>%
  distinct(sample_name) %>%
  nrow()

passedQC_count <- reprep_repool_summary %>%
  filter(status == "pass", !grepl("Control", sample_name)) %>%
  distinct(sample_name) %>%
  nrow()

total_samples <- repreps_count + repools_count + passedQC_count
if (total_samples > 0) {
  percentage_pass <- (passedQC_count / total_samples) * 100
  cat("Overall QC Summary:\n")
  cat("  Reprep:", repreps_count, "\n")
  cat("  Repool:", repools_count, "\n")
  cat("  Pass:", passedQC_count, "\n")
  cat("  Pass percentage:", round(percentage_pass, 1), "%\n")
}
```

:::

## Amplification Plate Maps

### Reads Heatmap

::: {.callout-tip collapse="true"}
## About this section
The reads heatmap provides a visual representation of read distribution across the plate, helping to identify inconsistencies in sequencing efficiency. This can highlight potential issues such as edge effects, batch effects, or pipetting errors that may impact data quality and interpretation.
:::

```{r}
plot_plate_with_feature <- function(quadrants_batch, sample_colours, batch_name, fill_param, scale_midpoint, scale_label) {
    plate_layout <- ggplot(quadrants_batch) +
      # Heatmap fill based on reads_per_reaction
      geom_tile(aes(x = Column, y = y, fill = !!rlang::parse_expr(fill_param)), color = NA, width = 0.95, height = 0.95) +
      # Separate tile layer for outlines
      geom_rect(data = quadrants_batch,
            aes(xmin = Column-0.45, xmax = Column+0.45, ymin = y-0.45, ymax = y+0.45,color=SampleType),fill = NA,linewidth=1.3)+
      # Facet by Batch and reaction
      facet_grid(reaction ~ Batch) +
      # Color gradient for heatmap
      scale_fill_gradient2(
        low = "black",
        mid = "darkorange4",
        high = "darkorange",
        midpoint = scale_midpoint,
        name = scale_label
      ) +
      # Outline colors based on SampleType
      scale_color_manual(values = sample_colours, na.translate = TRUE) +
      # Labels
      scale_y_continuous(breaks = quadrants_batch$y, labels = quadrants_batch$Row) +
      scale_x_continuous(breaks = 1:12, labels = as.character(1:12)) +
      xlab("Column") +
      ylab("Row") +
      # Theme adjustments
      theme_minimal() +
      theme(
        axis.text = element_text(size = 12),
        strip.text = element_text(size = 14, face = "bold"),
        panel.grid = element_blank(),
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        legend.key = element_rect(colour = "black")  # Ensure visible legend border
      ) + 
    coord_fixed(ratio = 0.75)
    return(plate_layout)
}

plot_plate_heat_maps <- function(sample_data, quadrants, sample_colours, fill_param, scale_midpoint, scale_label) {
    plate_maps = list()
    
    quadrants_by_reaction <- crossing(summary_samples %>% select(reaction) %>% distinct() , quadrants)
    merged_data <- merge(summary_samples, quadrants_by_reaction, by = c("Batch", "Row", "Column", "SampleType", "reaction"), all.y = TRUE)
    
    # Ensure NAs in SampleType are labeled
    merged_data$SampleType[is.na(merged_data$SampleType)] <- "empty"
    
    for(b in unique(merged_data$Batch)){
      merged_data_batch = merged_data %>%
        filter(Batch == b)
      plate_maps[[b]] = plot_plate_with_feature(merged_data_batch, sample_colours, b, fill_param, scale_midpoint, scale_label)
    }
    return(plate_maps)
}
```

```{r}
#| results: asis
#| echo: false
#| fig-column: screen-inset
plate_maps <- plot_plate_heat_maps(summary_samples, quadrants, sample_colours, "log10(reads_per_reaction + 0.1)", 2, "log10(reads_per_reaction)")
if (render_mode) {
  cat(create_tabsetOfGgplotObjects(plate_maps))
} else {
  for (plate in plate_maps) {
    print(plate)
  }
}
```

### Successful Amplification Heatmap
::: {.callout-tip collapse="true"}
## About this section
Here we visualise the success rate of amplification across the plate, allowing for the identification of poorly amplified regions or wells. This is useful for assessing the quality of the PCR process, ensuring that targets across all samples are adequately amplified, and helping to spot potential issues with specific wells or sample groups.
:::

```{r}
#| results: asis
#| echo: false
#| fig-column: screen-inset
plate_maps <- plot_plate_heat_maps(summary_samples, quadrants, sample_colours, "prop_good_loci", 0.5, "prop_good_loci")
if (render_mode) {
  cat(create_tabsetOfGgplotObjects(plate_maps))
} else {
  for (plate in plate_maps) {
    print(plate)
  }
}
```

## Save Output Files

```{r}
if (output_dir!="") {
    if (!dir.exists(output_dir)) {
        dir.create(output_dir, recursive = TRUE)
        cat("Created output directory:", output_dir, "\n")
      }
    reprep_repool_summary_path <- file.path(output_dir,reprep_repool_summary_path)
    polyclonal_information_path <- file.path(output_dir,polyclonal_information_path)
    neg_control_information_path <- file.path(output_dir, neg_control_information_path)
    samples_to_repool_path <- file.path(output_dir, samples_to_repool_path) 
    samples_to_reprep_path <- file.path(output_dir, samples_to_reprep_path)
    missing_samples_report_path <- file.path(output_dir, missing_samples_report_path)
    filtered_allele_data_path <- file.path(output_dir, filtered_allele_data_path)
    filtered_collapsed_allele_data_path <- file.path(output_dir, filtered_collapsed_allele_data_path)
    missing_targets_report_path <- file.path(output_dir, missing_targets_report_path)
    filtered_resmarker_table_path <- file.path(output_dir, filtered_resmarker_table_path)
    filtered_resmarker_microhaplotype_table_path <- file.path(output_dir,filtered_resmarker_microhaplotype_table_path)
}

# Create file paths 
sample_coverage_path <- file.path(results_dir,"sample_coverage.txt")

# Write reprep/repool summary
write.csv(reprep_repool_summary, reprep_repool_summary_path, row.names = FALSE)
cat("Saved:", reprep_repool_summary_path, "\n")

# Write polyclonal controls
write.csv(polyclonal_information, polyclonal_information_path, row.names = FALSE)
cat("Saved:", polyclonal_information_path, "\n")

# Write negative control contamination
write.csv(neg_control_information, neg_control_information_path, row.names = FALSE)
cat("Saved:", neg_control_information_path, "\n")

# Create separate CSV files for samples needing repool and reprep
samples_to_repool <- reprep_repool_summary %>%
  filter(status == "repool") %>%
  select(sample_name, Batch, reaction, reads_per_reaction, prop_good_loci, reason) %>%
  arrange(Batch, sample_name, reaction)

samples_to_reprep <- reprep_repool_summary %>%
  filter(status == "reprep") %>%
  select(sample_name, Batch, reaction, reads_per_reaction, prop_good_loci, reason) %>%
  arrange(Batch, sample_name, reaction)

write.csv(samples_to_repool, samples_to_repool_path, row.names = FALSE)
cat("Saved:", samples_to_repool_path, "(", nrow(samples_to_repool), "samples needing repool)\n")

write.csv(samples_to_reprep, samples_to_reprep_path, row.names = FALSE)
cat("Saved:", samples_to_reprep_path, "(", nrow(samples_to_reprep), "samples needing reprep)\n")

# Read in raw data with no other filtering applied
allele_data_for_output <- read.csv(allele_data_path, sep='\t')
collapsed_allele_data_for_output <- read.csv(collapsed_allele_data_path, sep='\t')
resmarker_table <- read.csv(resmarker_table_path, sep='\t')
resmarker_microhaplotype_table <- read.csv(resmarker_microhaplotype_table_path, sep='\t')

# Standardize SampleID
allele_data_for_output <- allele_data_for_output %>%
  mutate(sample_name = if (standardise_sample_name) sample_name %>%
           sub("_L001$", "", .) %>%
           word(., start = 1, end = -2, sep = "_") else SampleID)

collapsed_allele_data_for_output <- collapsed_allele_data_for_output %>%
  mutate(sample_name = if (standardise_sample_name) sample_name %>%
           sub("_L001$", "", .) %>%
           word(., start = 1, end = -2, sep = "_") else SampleID)
         
# Identify samples in manifest but not in allele data
samples_in_allele_data <- allele_data_for_output %>%
  distinct(sample_name) %>%
  pull(sample_name)

missing_samples <- manifest %>%
  filter(!(sample_name %in% samples_in_allele_data)) %>%
  filter(SampleType != 'negative') %>%
  select(sample_name, Batch, SampleType, Row, Column, Parasitemia) %>%
  arrange(Batch, sample_name)

write.csv(missing_samples, missing_samples_report_path, row.names = FALSE)
cat("Saved:", missing_samples_report_path, "(", nrow(missing_samples), "samples in manifest not found in allele data)\n")

# Calculate allele frequency
allele_data_for_output <- allele_data_for_output %>%
  group_by(sample_name, target_name) %>%
  mutate(AlleleFreq = reads / sum(reads)) %>%
  ungroup()

collapsed_allele_data_for_output <- collapsed_allele_data_for_output %>%
  group_by(sample_name, target_name) %>%
  mutate(AlleleFreq = reads / sum(reads)) %>%
  ungroup()

resmarker_table <- resmarker_table %>%
  group_by(sample_name, gene_id, gene, aa_position) %>%
  mutate(AlleleFreq = reads / sum(reads)) %>%
  ungroup()

group_cols <- c("sample_name", "gene_id", "gene", "target_name", "mhap_aa_positions", "ref_mhap")
resmarker_microhaplotype_table <- resmarker_microhaplotype_table %>%
  group_by(across(all_of(group_cols))) %>%
  mutate(AlleleFreq = reads / sum(reads)) %>%
  ungroup()
         
# Filter to minimum read and within sample allele frequency (if enabled)
if (filter_wsaf_final_allele_table) {
  allele_data_filtered <- allele_data_for_output %>%
    filter(reads > read_filter & AlleleFreq > af_filter)
  collapsed_allele_data_filtered <- collapsed_allele_data_for_output %>%
    filter(reads > read_filter & AlleleFreq > af_filter)
  resmarker_table_filtered <- resmarker_table %>%
    filter(reads > read_filter & AlleleFreq > af_filter)
  resmarker_microhaplotype_table_filtered <- resmarker_microhaplotype_table %>%
    filter(reads > read_filter & AlleleFreq > af_filter)
  filter_msg <- paste0("Reads > ", read_filter, " AND WSAF > ", af_filter)
} else {
  # Only filter by read count, not WSAF
  allele_data_filtered <- allele_data_for_output %>%
    filter(reads > read_filter)
  collapsed_allele_data_filtered <- collapsed_allele_data_for_output %>%
    filter(reads > read_filter)
  resmarker_table_filtered <- resmarker_table %>%
    filter(reads > read_filter)
  resmarker_microhaplotype_table_filtered <- resmarker_microhaplotype_table_filtered %>%
    filter(reads > read_filter)
  filter_msg <- paste0("Reads > ", read_filter, " (WSAF filter disabled)")
}

# Write out table
write.table(allele_data_filtered[ , !(names(allele_data_filtered) %in% c("AlleleFreq"))], filtered_allele_data_path, sep="\t", row.names = FALSE, quote = FALSE)
cat("Saved:", filtered_allele_data_path, "\n")
write.table(collapsed_allele_data_filtered[ , !(names(collapsed_allele_data_filtered) %in% c("AlleleFreq"))], filtered_collapsed_allele_data_path, sep="\t", row.names = FALSE, quote = FALSE)
cat("Saved:", filtered_collapsed_allele_data_path, "\n")
write.table(resmarker_table_filtered[ , !(names(resmarker_table_filtered) %in% c("AlleleFreq"))], filtered_resmarker_table_path, sep="\t", row.names = FALSE, quote = FALSE)
cat("Saved:", filtered_resmarker_table_path, "\n")
write.table(resmarker_microhaplotype_table_filtered[ , !(names(resmarker_microhaplotype_table_filtered) %in% c("AlleleFreq"))], filtered_resmarker_microhaplotype_table_path, sep="\t", row.names = FALSE, quote = FALSE)
cat("Saved:", filtered_resmarker_microhaplotype_table_path, "\n")
cat("  (Includes ALL targets, including those excluded from QC calculations)\n")
cat("  Filtered by:", filter_msg, "\n")
```

::: {.callout-note collapse="true"}
### Missing loci report 

```{r}
loci_not_in_allele_data <- panel_information %>%
    anti_join(allele_data, by = "target_name") %>%
    select(target_name, pool, chrom)

write.csv(loci_not_in_allele_data, missing_targets_report_path, row.names = FALSE)
cat("Saved:", missing_targets_report_path, "\n")

if (nrow(loci_not_in_allele_data) > 0) {
    cat("\nMissing Targets Report\n\n")
    cat("The following targets are present in the pool file but missing from the allele data:\n\n")

    missing_loci_table <- loci_not_in_allele_data %>%
      arrange(target_name) %>%
      mutate(No = row_number()) %>%
      select(No, everything())
    missing_loci_table |>
      kable() |>
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
  } else {
    cat("\nAll targets in pool file were found in the data.\n")
  }
```

:::

## QC Summary by Batch

::: {.callout-tip collapse="true"}
## About this section
This visualization shows the overall QC pass rate by batch, with stacked bars representing the percentage of samples passing, requiring repool, or requiring reprep. This helps identify batches with systematic issues.
:::

```{r}
#| fig-width: 10
#| fig-height: 6

# Calculate QC status counts by batch
qc_summary_by_batch <- reprep_repool_summary %>%
  filter(!grepl("Control", sample_name)) %>%  # Exclude controls
  distinct(sample_name, Batch, status) %>%
  group_by(Batch, status) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(Batch) %>%
  mutate(
    total = sum(count),
    percentage = (count / total) * 100
  ) %>%
  ungroup()

# Define colors for QC status
qc_colors <- c(
  "pass" = "Gray",
  "repool" = "skyblue",
  "reprep" = "orangered2"
)

# Create stacked bar plot
qc_plot <- ggplot(qc_summary_by_batch, aes(x = Batch, y = percentage, fill = status)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.3) +
  geom_text(
    aes(label = ifelse(percentage > 5, paste0(round(percentage, 1), "%"), "")),
    position = position_stack(vjust = 0.5),
    size = 5,
    fontface = "bold",
    color = "black"
  ) +
  scale_fill_manual(
    values = qc_colors,
    name = "QC Status",
    breaks = c("pass", "repool", "reprep"),
    labels = c("Pass", "Repool", "Reprep")
  ) +
  labs(
    title = "QC Status by Batch",
    x = "Batch",
    y = "Percentage of Samples (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.position = "top",
    panel.grid.major.x = element_blank()
  ) +
  ylim(0, 100)

print(qc_plot)
```

```{r}
# Print summary table
qc_summary_table <- reprep_repool_summary %>%
  filter(!grepl("Control", sample_name)) %>%
  distinct(sample_name, Batch, status) %>%
  group_by(Batch, status) %>%
  summarize(count = n(), .groups = "drop") %>%
  # Ensure all status levels exist for each batch
  complete(Batch, status = c("pass", "repool", "reprep"), fill = list(count = 0)) %>%
  pivot_wider(names_from = status, values_from = count, values_fill = 0) %>%
  mutate(
    total = pass + repool + reprep,
    pass_rate = round(100 * pass / total, 1)
  ) %>%
  select(Batch, pass, repool, reprep, total, pass_rate)

qc_summary_table %>%
  kable(col.names = c("Batch", "Pass", "Repool", "Reprep", "Total", "Pass Rate (%)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Summary

::: {.callout-important collapse="false"}
## Report Summary

This unified QC report has processed data and generated the following standardized outputs:

1. **`r reprep_repool_summary_path`** - Combined reprep/repool classification for all samples
2. **`r samples_to_repool_path`** - Samples requiring repool
3. **`r samples_to_reprep_path`** - Samples requiring reprep
4. **`r polyclonal_information_path`** - Polyclonality information for positive controls
5. **`r neg_control_information_path`** - Contamination detected in negative controls
6. **`r filtered_allele_data_path`** - Filtered allele data meeting read/AF thresholds (includes all targets)
7. **`r missing_targets_report_path`** - Targets in pool file but absent from data (if applicable)
8. **`r missing_samples_report_path`** - Samples in manifest but not found in allele data

### Key Points

- **Targets filtered for QC**: `r length(excluded_loci)` targets were excluded from QC calculations
- **QC metrics**: Reprep/repool classifications based on `r n_distinct(amplicon_coverage_qc$target_name)` QC-relevant targets
- **Final output**: `allele_data_filtered.txt` includes ALL targets (both QC and excluded) meeting thresholds
- **Pass rate**: `r if(exists("percentage_pass")) paste0(round(percentage_pass, 1), "%") else "N/A"`
:::


